---
title: "Comparing Triplet Embeddings to Alternative Representations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing Triplet Embeddings to Alternative Representations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Overview

A triplet-based embedding is one way to estimate the similarity structure among a set of items, but it is rarely the only way. You might also have:

- A **language- or vision-model embedding** of the same items (e.g. from a HuggingFace model) — typically much higher-dimensional than a triplet embedding
- A **categorical label** for each item (a stimulus feature, a taxonomy category) that a good embedding ought to encode
- A **verbal-fluency task** (e.g. "name as many animals as you can") from which you can estimate similarity via the sequential structure of what people say

This vignette covers three tools for relating a triplet embedding to each of these:

| Approach | Question it answers | Key functions |
|---|---|---|
| Procrustes alignment + dimensionality ceilings | How well does this embedding's *shape* align with a (possibly much higher-dimensional) alternative? | `get.rep.dist()`, `procrustes_rank_ceiling()`, `procrustes_spectral_ceiling()` |
| Classifier performance | Does this embedding encode a known categorical label? | `repeated_stratified_logistic_cv()`, `repeated_stratified_multinomial_cv()` |
| Successor representations from verbal fluency | How does similarity estimated from a completely different task (free naming) compare to triplet-based similarity? | `successor_matrix()`, `hellinger_dist()` |

All three sections build on real example data bundled with the package — Shaver et al. (1987) emotion words with a matched language-model embedding, the icon stimulus set, and an animal verbal-fluency dataset — so every number shown below is a real, reproducible result rather than a synthetic illustration.

```{r load}
library(tripletTools)
```

---

## Comparing embeddings via Procrustes alignment

### The naive approach, and a pitfall to avoid

The obvious way to compare two embeddings of the same items is a Procrustes fit: find the rotation, reflection, and scaling of one configuration that best superimposes it on the other, and report how good the best-possible fit is. `get.rep.dist()` already does this for pairs of embeddings in this package (it's normally used to compare embeddings across participants):

```{r naive-compare}
sdist <- get.rep.dist(list(triplet = emotion_triplet_embedding, bge = emotion_bge_embedding))
observed_corr <- 1 - sdist[1, 2]   # get.rep.dist() returns a *distance*; 1 - distance is the correlation
observed_corr
#> 0.313
```

(`get.rep.dist()`'s default distance is `1 - sqrt(1 - ss)`, the Procrustes equivalent of a correlation — see its own documentation. Its output matrix doesn't carry the input list's names, hence indexing by position.)

**One thing worth flagging before you reach for this on your own data:** it's tempting, when comparing two embeddings, to first convert each to an item-by-item distance matrix and Procrustes-align *those* — especially if you want a method that doesn't care about the two embeddings' raw dimensionality. Don't do this. Euclidean distance is a *nonlinear* function of coordinates, so a raw pairwise distance matrix computed from a genuinely low-rank configuration is generally **not** itself low-rank — a truly 3-dimensional configuration's distance matrix will typically have rank close to `n - 1`, not 3. Feeding that distance matrix into rank/ceiling analysis (below) as if it were a coordinate matrix will badly overstate the configuration's effective dimensionality. Just Procrustes-align the raw coordinate matrices directly, as above — `vegan::procrustes()` (which `get.rep.dist()` calls internally) automatically zero-pads the lower-dimensional one, so mismatched dimensionality is handled correctly with no extra steps.

### Why dimensionality mismatch still matters

`emotion_triplet_embedding` is 4-dimensional. `emotion_bge_embedding` — a [BAAI/bge-m3](https://huggingface.co/BAAI/bge-m3) language-model embedding of the same 213 words — is natively 1024-dimensional (bundled here rotated onto its own lossless top-212 principal components, since with only 213 items its effective rank can't exceed 212 regardless of the model's native output width; see `?emotion_bge_embedding`). A correlation of 0.313 between a 4-dimensional and an effectively-212-dimensional configuration is hard to interpret in isolation: is 0.313 a *good* alignment, given that a 4-dimensional space can only ever capture a slice of whatever a 212-dimensional space represents? Or is it a *poor* one, reflecting real representational mismatch beyond what dimensionality alone would explain?

`procrustes_rank_ceiling()` and `procrustes_spectral_ceiling()` answer this by computing an upper bound on the achievable correlation, so you can judge the observed value against what was actually attainable rather than against an abstract 0-to-1 scale.

```{r ceilings}
procrustes_rank_ceiling(emotion_bge_embedding, k = 4)
#> 0.436

procrustes_spectral_ceiling(emotion_triplet_embedding, emotion_bge_embedding)
#> 0.418
```

- `procrustes_rank_ceiling(target, k)` asks: of all possible rank-`k` configurations, what's the best correlation *any* of them could achieve against `target`? This is a property of `target` alone (via its singular-value spectrum), independent of any particular candidate embedding.
- `procrustes_spectral_ceiling(candidate, target)` asks a related but sharper question: given `candidate`'s *actual* singular-value spectrum (not just its nominal rank), what's the best correlation it could achieve against `target` under optimal rotation? `procrustes_rank_ceiling(target, k)` is the special case of this where `candidate` is `target`'s own best rank-`k` approximation.

Both are exact, provable bounds (via the Eckart–Young/Ky Fan theorem and von Neumann's trace inequality respectively) under the same symmetric, scaled Procrustes statistic `get.rep.dist()` uses — not heuristics.

### Interpreting the result

Observed correlation (0.313) against the two ceilings (0.436, 0.418) means the triplet embedding captures roughly **72–75% of what was achievable** at its dimensionality. That's a reassuring result: most of the shortfall from a "perfect" correlation of 1 is explained by the sheer dimensionality gap between a 4D and an effectively-212D space, not by the triplet embedding capturing fundamentally different structure. Without the ceiling, 0.313 alone doesn't tell you that — it could just as easily have reflected a triplet embedding that captures almost none of the achievable structure.

---

## Evaluating embeddings via classifier performance

### The idea

If an embedding faithfully captures a property of the items, a classifier trained on the embedding's coordinates should be able to predict that property from held-out items. `repeated_stratified_logistic_cv()` (binary labels) and `repeated_stratified_multinomial_cv()` (more than two classes) estimate this via repeated stratified k-fold cross-validation, reporting AUC and related metrics along with per-fold diagnostics for catching unreliable fits.

### Binary labels: individual differences in what gets encoded

The icon stimulus set's item codes encode several features directly (see `?icon_items`): position 1 is face/place, position 2 is day/night. Both apply to all 32 items, so both make clean binary labels. Rather than asking whether the *group* embedding encodes these features, it's more interesting to ask whether *individual* participants' embeddings do — different people may attend to different stimulus properties when judging similarity.

```{r binary-setup}
items <- icon_items$item
faceplace <- ifelse(substr(items, 1, 1) == "f", "face", "place")
daynight  <- ifelse(substr(items, 2, 2) == "d", "day", "night")
names(faceplace) <- items
names(daynight)  <- items

run_feature <- function(label) {
  sapply(icon_emb_ind, function(emb) {
    X <- emb[items, ]
    y <- label[items]
    cv <- repeated_stratified_logistic_cv(X, y, repetitions = 100, folds = 4, seed = 1)
    cv$summary$mean[cv$summary$metric == "auc"]
  })
}

data.frame(
  worker_id  = names(icon_emb_ind),
  face_place = run_feature(faceplace),
  day_night  = run_feature(daynight),
  row.names  = NULL
)
```

```
#>   worker_id face_place day_night
#> 1  3n7ggxph      1.000     0.307
#> 2  b5wma4no      0.998     0.566
#> 3  d8mmm1qn      1.000     0.361
#> 4  jn7bbjc0      1.000     0.536
#> 5  pbby694o      0.616     1.000
#> 6  sc2xbd6w      0.808     1.000
```

Four participants (`3n7ggxph`, `b5wma4no`, `d8mmm1qn`, `jn7bbjc0`) encode face/place almost perfectly (AUC 0.998–1.000) but only weakly for day/night (0.31–0.57, closer to the chance level of 0.5 than to 1). The other two (`pbby694o`, `sc2xbd6w`) show the reverse: perfect day/night discrimination but only moderate face/place discrimination. This is a genuine crossover, not just noise on one feature or the other — the *same* participants swap roles depending on which property you ask about, which is a much stronger illustration of individual differences in representation than either feature alone would give.

With only 32 items split across 4 folds, some fits in this example will be flagged `unstable_fit` in `cv$diagnostics` (most often the near-perfect-AUC fits, for the same quasi-complete-separation reason discussed below) — small item sets make this common. Inspect `cv$diagnostics` before trusting any single fold's coefficients in isolation.

### More than two classes: predicting emotion category from embedding coordinates

`emotion_categories` gives each of the 213 Shaver et al. (1987) emotion words a basic-emotion category (Love, Joy, Surprise, Anger, Sadness, Fear), plus `"Absent"` for words Shaver et al. didn't end up assigning to a category. Before classifying, drop `"Absent"` — and also `"Surprise"`, which only has 3 words, too few for any fold to give a stable estimate:

```{r multinomial-setup}
keep <- !(emotion_categories$category %in% c("Absent", "Surprise"))
words <- rownames(emotion_categories)[keep]

X <- emotion_triplet_embedding[words, ]
y <- droplevels(emotion_categories[words, "category"])
table(y)
#>   Anger    Fear     Joy    Love Sadness
#>      29      17      33      16      37
```

```{r multinomial-fit}
cv <- repeated_stratified_multinomial_cv(X, y, repetitions = 20, folds = 3, seed = 1)
cv$summary
```

```
#>              metric   mean     sd
#>            accuracy  0.698  0.021
#>   balanced_accuracy  0.638  0.024
#>           precision  0.650  0.028
#>                  f1  0.639  0.025
#>                 auc  0.924  0.013
```

An overall AUC of 0.924 looks strong, but the macro-averaged metrics above collapse across all 5 classes — `per_class_summary` shows they don't all fare the same:

```{r multinomial-perclass}
cv$per_class_summary[, c("class", "n", "mean_sensitivity", "mean_precision", "mean_f1")]
```

```
#>     class  n mean_sensitivity mean_precision mean_f1
#> 1   Anger 29            0.741          0.675   0.706
#> 2    Fear 17            0.529          0.622   0.568
#> 3     Joy 33            0.761          0.684   0.719
#> 4    Love 16            0.297          0.408   0.341
#> 5 Sadness 37            0.861          0.859   0.859
```

`Sadness` is predicted well (0.86 sensitivity); `Love` is predicted poorly (0.30) — a pattern the macro-averaged `balanced_accuracy` alone would hide entirely.

### A worked example in calibrating `coefficient_warning`

Before trusting the fit above, it's worth checking `cv$diagnostics` for signs of instability — and here, the default threshold turns out to need recalibrating for this problem:

```{r multinomial-diagnostics-default}
summary(cv$diagnostics$maximum_absolute_coefficient)
#>    Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
#>   11.57   18.11   25.50  131.68   35.18 2291.16

sum(cv$diagnostics$unstable_fit)
#> 60   # out of 60 -- every single fold
```

Every fold is flagged, which makes the diagnostic useless as a discriminating signal here. This isn't a bug — it's `coefficient_warning`'s default of `10` being tuned for `repeated_stratified_logistic_cv()`'s 2-class case, where it's a reasonable threshold. With 5 classes fit from only 4 predictor dimensions, multinomial coefficients naturally run larger even in well-behaved fits (the median here, 25.5, is itself well above 10), so the same threshold doesn't transfer. The right fix isn't to trust a hard-coded default blindly, but to look at the empirical distribution of `maximum_absolute_coefficient` across folds — as above — and pick a threshold that separates the typical range from real outliers.

The distribution above has a clear typical band (roughly 10–50) and a small outlier tail (6 fits above 100, one at 2291). Recalibrating:

```{r multinomial-recalibrated}
cv2 <- repeated_stratified_multinomial_cv(X, y, repetitions = 20, folds = 3, seed = 1,
                                           coefficient_warning = 50)
sum(cv2$diagnostics$unstable_fit)
#> 9   # out of 60 -- now a genuinely discriminating signal
```

`cv2$summary` and `cv2$per_class_summary` are unchanged from `cv`'s — `coefficient_warning` only affects the diagnostic flag, not the fit or predictions. The one fold reaching a coefficient of 2291 is worth a closer look on its own merits: despite `nnet::multinom` reporting convergence and a non-degenerate training accuracy (0.79, not the ~1.0 you'd expect from *complete* separation), 301 of its 430 training predicted-probability cells were within numerical tolerance of 0 or 1 — a subtler, partial-separation pattern for a subset of predictions that a simple "is training accuracy near 100%" check wouldn't have caught, but the coefficient-magnitude diagnostic did.

---

## Estimating similarity from verbal-fluency data

### Successor representations

A verbal-fluency task — "name as many animals as you can" — produces sequential lists rather than similarity judgments directly, but sequential structure still carries information about similarity: items that follow one another within a list tend to be related. Counting how often each word follows each other word across many lists gives a directed, weighted adjacency graph. `successor_matrix()` turns that graph into a **Successor Representation**: for each pair of items, the discounted expected number of times a random walk from one would visit the other, summed over an infinite horizon.

```{r successor-basic}
sr <- successor_matrix(animal_adjacency_matrix, gamma = 0.5)
dim(sr$successor)
#> [1] 295 295
```

`gamma` (0 ≤ gamma < 1) controls the time horizon: near 0, the representation reflects only immediate neighbors; as it approaches 1, it increasingly reflects long-range graph structure. There's no universally "correct" value — see the sensitivity check at the end of this section.

### Restricting to matching items — after, not before, computing the successor representation

`animal_adjacency_matrix` has 295 words (anything produced by at least 5 participants in the underlying fluency study), while `animal_triplet_embedding` has only the 213 that were part of the curated triplet stimulus set. The other 82 are superordinate terms, synonyms, and other species that came up in people's lists but weren't part of the triplet study.

**Compute `successor_matrix()` on the full 295-word graph, and only restrict to the 213 matching items afterward:**

```{r successor-restrict}
target_items <- rownames(animal_triplet_embedding)
S <- sr$successor[target_items, target_items]   # subset AFTER computing, not before
```

Restricting the *adjacency matrix* to 213 words before computing the successor representation would discard genuine multi-hop associative structure that happens to route through a non-target word — e.g. a real bear → cub → lion pathway would vanish entirely if "cub" were dropped before the random walk was ever computed, even though bear and lion are both target items. Computing on the full graph first preserves that structure; only the final read-out is restricted.

### From successor representations to distances: `hellinger_dist()`

`S`'s rows aren't directly comparable to `animal_triplet_embedding`'s coordinates — they're profiles over 213 other items, not embedding coordinates in the usual sense. `hellinger_dist()` treats each row as a probability profile (renormalizing it to sum to 1) and returns the pairwise Hellinger distance between rows:

```{r hellinger-basic}
D_successor <- hellinger_dist(S)
```

Here's a fact worth knowing before reaching for classical MDS on `D_successor` to get something Procrustes-comparable: **Hellinger distance is not merely well-approximated by a Euclidean distance — it *is* one, exactly, by definition.** `hellinger_dist()` computes `dist(sqrt(P)) / sqrt(2)` internally, which means `D_successor` is *exactly* the (scaled) Euclidean distance matrix of `sqrt(row-normalized S)`. There's no approximation risk here the way there might be for an arbitrary dissimilarity matrix — classical MDS on `D_successor` would recover `sqrt(row-normalized S)` (up to rotation) with zero distortion, so it's simpler to just compute that directly:

```{r hellinger-coords}
P <- S
P[P < 0] <- 0
P <- P / rowSums(P)
Q <- sqrt(P)              # exact coordinates implied by hellinger_dist(S), up to rotation
rownames(Q) <- target_items
```

`Q` is now a genuine coordinate matrix, directly comparable to `animal_triplet_embedding` via the same Procrustes/ceiling tools from the first section of this vignette.

### Comparing to the triplet embedding

```{r successor-compare}
sdist <- get.rep.dist(list(triplet = animal_triplet_embedding, successor = Q))
observed_corr <- 1 - sdist[1, 2]
observed_corr
#> 0.282

procrustes_rank_ceiling(Q, k = 6)
#> 0.410

procrustes_spectral_ceiling(animal_triplet_embedding, Q)
#> 0.367
```

As with the language-model comparison earlier, `Q` (like `S` itself) is naturally high-rank — a 6-dimensional triplet embedding can only ever capture part of its structure, which is exactly what the ceiling quantifies (0.410 at k = 6, well short of 1). The observed correlation (0.282) falls short of even that reduced ceiling, capturing about 69% of what was achievable at rank 6 — a similar shortfall to the language-model comparison, though the underlying "alternative representation" here is estimated from a completely different task (free naming, not a language model) and depends on a parameter (`gamma`) that has no counterpart in the earlier comparison.

### Sensitivity to gamma

Because there's no principled default for `gamma`, it's worth checking whether conclusions are sensitive to it before reporting a result at a single value:

```{r gamma-sweep}
gammas <- c(0.3, 0.5, 0.7, 0.9)

sweep <- do.call(rbind, lapply(gammas, function(g) {
  sr_g <- successor_matrix(animal_adjacency_matrix, gamma = g)
  S_g <- sr_g$successor[target_items, target_items]
  P_g <- S_g; P_g[P_g < 0] <- 0; P_g <- P_g / rowSums(P_g)
  Q_g <- sqrt(P_g); rownames(Q_g) <- target_items

  sdist_g <- get.rep.dist(list(triplet = animal_triplet_embedding, successor = Q_g))
  data.frame(
    gamma          = g,
    observed_corr  = 1 - sdist_g[1, 2],
    rank_ceiling   = procrustes_rank_ceiling(Q_g, k = ncol(animal_triplet_embedding)),
    spectral_ceiling = procrustes_spectral_ceiling(animal_triplet_embedding, Q_g)
  )
}))
sweep
```

```
#>   gamma observed_corr rank_ceiling spectral_ceiling
#> 1   0.3         0.249        0.356            0.315
#> 2   0.5         0.282        0.410            0.367
#> 3   0.7         0.308        0.451            0.409
#> 4   0.9         0.316        0.466            0.429
```

Alignment with the triplet embedding rises with `gamma` here — but treat that as a demonstration that the choice matters and is worth checking on your own data, not as a general claim about what `gamma` "should" be for verbal-fluency data. The direction and size of this kind of effect isn't obvious in advance (a different comparison space, e.g. a language-model embedding of the same words, need not move the same way with `gamma`), which is itself the point: a successor representation's geometry is not a fixed thing, and the discount parameter is worth exploring rather than fixed at a conventional default without checking.

---

## Summary

| Question | Tool |
|---|---|
| Does this embedding's overall shape align with an alternative, accounting for a dimensionality mismatch? | `get.rep.dist()` + `procrustes_rank_ceiling()`/`procrustes_spectral_ceiling()` |
| Does this embedding encode a known 2-level label? | `repeated_stratified_logistic_cv()` |
| Does this embedding encode a known >2-level label? | `repeated_stratified_multinomial_cv()` |
| How does verbal-fluency-based similarity compare to this embedding? | `successor_matrix()` + `hellinger_dist()`, then the Procrustes tools above |

All three tools share a common thread: a raw number in isolation (a correlation, an AUC, a distance) is hard to interpret without something to compare it to — a ceiling, a chance level, a per-class breakdown, a sensitivity check across a hyperparameter. Each section above pairs the raw computation with that context.
