---
title: "Computing Triplet Embeddings"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Computing Triplet Embeddings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Overview

This vignette explains how to compute similarity embeddings from triplet judgment data collected with this package. Embeddings are computed by a Python backend that uses PyTorch to optimize the crowd kernel (CKL) noise model. From R you interact with this backend through three functions:

| Function | What it does |
|---|---|
| `setup_python_env()` | One-time installation of the Python environment |
| `estimate_dimensionality()` | Fits embeddings across a range of dimensions to help choose `d` |
| `run_group_embedding_from_list()` | Trains a single embedding on all participants' data combined |
| `run_embeddings_from_list()` | Trains one embedding per participant plus a group embedding |

All functions accept data in the named-list format used throughout `tripletTools` (e.g. `icon_triplets`).

```{r load}
library(tripletTools)
```

---

## Step 1: One-time Python environment setup

The embedding pipeline runs inside a self-contained conda environment. You only need to do this once after installing the package. It installs PyTorch, NumPy, pandas, scikit-learn, scipy, skorch, and setuptools into an environment called `"triplet-embeddings"`.

```{r setup-env}
setup_python_env()
```

PyTorch is a large download (300–800 MB for a CPU build), so the first run may take several minutes. On future R sessions you do **not** need to call `setup_python_env()` again — the environment is detected and activated automatically when you load the package.

If conda is not found on your system, `setup_python_env()` will stop with a message directing you to the Miniconda installation page.

### If conda is installed but not found

On Windows, conda is sometimes installed without being added to the system PATH, which means R cannot locate it automatically. If you see a "Conda was not found" error despite having Miniconda or Anaconda installed, find the conda executable and tell reticulate where it is:

```{r conda-path}
# Common locations — run this to find yours:
possible_paths <- c(
  file.path(Sys.getenv("USERPROFILE"), "miniconda3", "Scripts", "conda.exe"),
  file.path(Sys.getenv("USERPROFILE"), "anaconda3", "Scripts", "conda.exe"),
  "C:/ProgramData/miniconda3/Scripts/conda.exe",
  "C:/ProgramData/anaconda3/Scripts/conda.exe",
  "C:/miniconda3/Scripts/conda.exe"
)
possible_paths[file.exists(possible_paths)]
```

Once you have the path, set it for the current session:

```{r conda-path-set}
Sys.setenv(RETICULATE_CONDA = "C:/Users/you/miniconda3/Scripts/conda.exe")
```

To avoid having to do this every session, add the line to your `.Renviron` file so it is set automatically on startup:

```{r renviron}
usethis::edit_r_environ()
# Add this line to the file that opens, then save and restart RStudio:
# RETICULATE_CONDA=C:/Users/you/miniconda3/Scripts/conda.exe
```

### GPU / CUDA support

If your machine has an NVIDIA GPU, you can install a CUDA-enabled build of PyTorch by passing `cuda_version`. First check your maximum supported CUDA version by running `nvidia-smi` in a terminal (look for "CUDA Version" in the top-right corner), then pass any version at or below that number:

```{r setup-env-cuda}
setup_python_env(cuda_version = "12.4")  # common values: "11.8", "12.1", "12.4"
```

On Windows, CUDA libraries are bundled directly into the PyTorch build from the pytorch conda channel, so the version number selects the closest available build rather than an exact match. On Linux and macOS the CUDA runtime is installed as a separate package from the nvidia conda channel. CUDA-enabled builds are substantially larger (~1.5–2 GB).

---

## Step 2: Computing a group embedding

`run_group_embedding_from_list()` pools the judgments from all participants and trains a single embedding that captures the shared similarity structure across the group. This is the fastest option and is a natural starting point before computing individual embeddings.

```{r group-embedding}
grp <- run_group_embedding_from_list(
  triplet_list = icon_triplets,
  d            = 3L,       # number of dimensions
  max_epochs   = 50000L
)
```

The function prints training progress every 100 epochs, showing train loss, test loss, train accuracy, and test accuracy. Training stops early if the test loss has not improved meaningfully for 10,000 consecutive epochs.

### What is returned

`run_group_embedding_from_list()` returns a named list:

```{r group-result-str}
str(grp)
#> List of 3
#>  $ embedding: num [1:32, 1:3] ...
#>   ..- attr(*, "dimnames")=List of 2
#>   .. ..$ : chr [1:32] "bdbwb" "bdbwb2" ...   # item names as row names
#>   .. ..$ : chr [1:3] "dim_0" "dim_1" "dim_2"
#>  $ loss     : num 0.412
#>  $ history  :'data.frame': 847 obs. of 5 variables:
#>   ..$ epoch     : int [1:847] 0 1 2 ...
#>   ..$ train_loss: num [1:847] ...
#>   ..$ test_loss : num [1:847] ...
#>   ..$ train_acc : num [1:847] ...
#>   ..$ test_acc  : num [1:847] ...
```

- `embedding` — a matrix with one row per stimulus (item names as row names) and `d` columns (`dim_0`, `dim_1`, …).
- `loss` — the best test loss achieved during training.
- `history` — a data frame recording train/test loss and accuracy at each epoch, useful for diagnosing convergence.

### Inspecting the result

```{r group-inspect}
# First few rows of the embedding
head(grp$embedding)

# Best test loss
cat("Best test loss:", round(grp$loss, 3), "\n")

# Did training converge? Plot the test loss curve.
plot(grp$history$epoch, grp$history$test_loss,
     type = "l", col = "steelblue",
     xlab = "Epoch", ylab = "Test loss",
     main = "Group embedding – training curve")
```

### Visualizing the embedding

Once you have the group embedding, you can use `plot_pics()` to position stimulus images at their coordinates in the learned similarity space (first two dimensions shown):

```{r group-plot, fig.width = 8, fig.height = 8}
plot_pics(grp$embedding[, 1:2], icon_pics,
          psize = 0.04,
          xlab  = "Dimension 1",
          ylab  = "Dimension 2",
          main  = "Group embedding")
```

Items that appear close together were frequently judged as similar by participants overall.

The group embedding can also be used with the full set of `tripletTools` analysis functions. For example, to evaluate how well it predicts held-out judgments for participant 1:

```{r group-hoacc}
acc <- get.hoacc(grp$embedding, icon_triplets[[1]])
cat("Hold-out accuracy (participant 1):", round(acc, 3), "\n")
```

---

## Step 3: Computing individual embeddings

`run_embeddings_from_list()` trains a separate embedding for each participant and then trains an additional group embedding on all participants' data combined.  This takes longer but lets you study individual differences in how participants represent the stimulus set.

```{r ind-embeddings}
results <- run_embeddings_from_list(
  triplet_list = icon_triplets,
  output_dir   = "embeddings_output",   # CSVs are also saved here
  d            = 3L,
  max_epochs   = 50000L
)
```

Output CSV files are written to `output_dir` as a side-effect:

| File | Contents |
|---|---|
| `embeddings.csv` | All per-participant and group embeddings concatenated |
| `embeddings_group.csv` | Group embedding only |
| `model_history.csv` | Training diagnostics for each participant and the group |

### What is returned

`run_embeddings_from_list()` returns a named list:

```{r ind-result-str}
str(results, max.level = 2)
#> List of 3
#>  $ individual:List of 6
#>   ..$ 3n7ggxph: num [1:32, 1:3] ...   # one matrix per participant
#>   ..$ b5wma4no: num [1:32, 1:3] ...
#>   ..$ d8mmm1qn: num [1:32, 1:3] ...
#>   ..$ jn7bbjc0: num [1:32, 1:3] ...
#>   ..$ pbby694o: num [1:32, 1:3] ...
#>   ..$ sc2xbd6w: num [1:32, 1:3] ...
#>  $ group     : num [1:32, 1:3] ...    # group embedding matrix
#>  $ history   :'data.frame': 7 obs. of 6 variables:
#>   ..$ worker_id              : chr [1:7] "3n7ggxph" ... "group"
#>   ..$ lowest_loss            : num [1:7] ...
#>   ..$ epoch                  : int [1:7] ...
#>   ..$ counter_from_last_update: int [1:7] ...
#>   ..$ n_train_triplets       : int [1:7] ...
#>   ..$ n_test_triplets        : int [1:7] ...
```

- `individual` — a named list of matrices, one per participant, in the same format as `grp$embedding` above.  Element names match the `worker_id` labels in your data.
- `group` — a single matrix containing the group embedding.
- `history` — a data frame with one row per participant plus one row for the group, summarising training diagnostics.

### Inspecting training quality

The `history` data frame gives a quick overview of how well each participant's model converged:

```{r history-inspect}
results$history[, c("worker_id", "lowest_loss", "epoch", "n_train_triplets")]
```

Participants who stopped early (small `epoch`) and achieved a low `lowest_loss` are well-modelled. If a participant's `lowest_loss` is much higher than the others, their data may be noisier or they may have fewer trials.

### Working with individual embeddings

The individual embeddings are stored in `results$individual` and have the same structure as the group embedding: a matrix with item row names and columns `dim_0`, `dim_1`, ….  This means they are compatible with all existing `tripletTools` functions.

```{r ind-use}
# Hold-out prediction accuracy for each participant using their own embedding
sapply(names(results$individual), function(wid) {
  get.hoacc(results$individual[[wid]], icon_triplets[[wid]])
})

# Prediction matrix: how well does each participant's embedding predict
# every other participant's held-out judgments?
pmat <- get.prediction.matrix(results$individual, icon_triplets)

# Do participants predict their own data better than others predict it?
cat("Mean self-prediction accuracy:  ", round(mean(diag(pmat)), 3), "\n")
cat("Mean other-prediction accuracy: ", round(mean(pmat[row(pmat) != col(pmat)]), 3), "\n")
```

### Visualizing individual embeddings

```{r ind-plot, fig.width = 8, fig.height = 8}
emb1 <- results$individual[[1]]

plot_pics(emb1[, 1:2], icon_pics,
          psize = 0.04,
          xlab  = "Dimension 1",
          ylab  = "Dimension 2",
          main  = paste("Individual embedding –", names(results$individual)[1]))
```

---

## Choosing between the two functions

| | `run_group_embedding_from_list()` | `run_embeddings_from_list()` |
|---|---|---|
| **Speed** | Fast — one model | Slow — one model per participant + group |
| **Output** | Single group embedding | Per-participant + group embeddings |
| **Best for** | Quick summary of shared structure; pilot analyses | Studying individual differences; full analysis |
| **Saved files** | None | Three CSVs in `output_dir` |

If you only need the group embedding, prefer `run_group_embedding_from_list()`. If you want to study whether participants differ in their representations — using tools like `get.prediction.matrix()`, `get.rep.dist()`, or `pacc.by.cluster()` — run `run_embeddings_from_list()` to obtain individual embeddings as well.

---

## Controlling training

Both functions accept the same set of hyperparameters:

| Parameter | Default | Meaning |
|---|---|---|
| `d` | `5` | Number of embedding dimensions |
| `max_epochs` | `50000` | Maximum training epochs |
| `tolerance` | `1e-4` | Minimum loss improvement to reset the early-stopping counter |
| `tol_window` | `10000` | Epochs without improvement before stopping |
| `seed` | `222` | Random seed for reproducibility |

For exploratory analyses, the defaults are a reasonable starting point. For a final analysis you may want to increase `max_epochs` and `tol_window` to allow longer convergence, especially with large datasets or high `d`.

To inspect the training curve after fitting:

```{r training-curve}
plot(grp$history$epoch, grp$history$test_loss,
     type = "l", col = "steelblue", lwd = 2,
     xlab = "Epoch", ylab = "Test loss",
     main = "Convergence check")
lines(grp$history$epoch, grp$history$train_loss,
      col = "tomato", lwd = 2, lty = 2)
legend("topright",
       legend = c("Test", "Train"),
       col    = c("steelblue", "tomato"),
       lty    = c(1, 2), lwd = 2)
```

A well-converged model shows test loss flattening well before `max_epochs`. If the loss is still falling at the end, increase `max_epochs`.

---

## Spherical embeddings

By default, embeddings are computed freely in Euclidean space — items can end up anywhere in \(\mathbb{R}^d\). Sometimes you have reason to believe the true structure is instead **circular or spherical**: a periodic variable like hue, phase, or time-of-day, where "wrapping around" is part of the representation rather than an artifact to be fit away. `train_embedding()` and every function built on it accept a `geometry` argument for exactly this case.

```{r sphere-basic}
grp_circle <- run_group_embedding_from_list(
  triplet_list = icon_triplets,
  d            = 2L,          # d = 2 means the *ambient* dimension —
  geometry     = "sphere",    # a 2-sphere embedded in R^2 is a circle
  radius       = 1,
  max_epochs   = 50000L
)
```

`d` always refers to the ambient dimension the sphere lives in, matching the convention used everywhere else in the package. A circle is the `d = 2` case; `d = 3` gives an ordinary sphere; higher `d` gives a hypersphere. `radius` sets the sphere's radius (default `1`) — mostly cosmetic, since it just rescales the recovered coordinates.

### Why fitting takes two stages

Constraining every point to a sphere's surface removes one degree of freedom per item relative to a free Euclidean fit. That turns out to matter a lot for optimization: starting from a random arrangement *on* the sphere, gradient descent has too little room to maneuver and reliably gets stuck reproducing close to chance-level accuracy, unable to escape a bad ordering of the items around the sphere.

To avoid this, `geometry = "sphere"` fits in two stages automatically:

1. A free Euclidean embedding is trained first (same `d`, same `max_epochs`/`tolerance`/`tol_window`), which easily finds a good arrangement because it isn't constrained.
2. That solution is projected onto the sphere and used as the starting point for a second fit, now constrained to stay on the sphere's surface.

You'll see both stages' progress printed in the console, labelled `[spherical warm start: fitting free Euclidean embedding]` and `[fitting constrained spherical embedding]`. This roughly **doubles training time** relative to `geometry = "euclidean"` — budget for it when setting `max_epochs` for a spherical fit on a large dataset.

### Interpreting a circular (d = 2) embedding

For `d = 2`, each item's angle around the circle is often more interpretable than its raw `(x, y)` coordinates:

```{r sphere-angles}
angles <- atan2(grp_circle$embedding[, 2], grp_circle$embedding[, 1])
names(angles) <- rownames(grp_circle$embedding)
sort(angles)
```

As with any embedding, the absolute rotation and reflection are arbitrary (not identifiable from triplet judgments alone) — only the relative circular *order* and *spacing* of items are meaningful. Plotting the fit with an equal aspect ratio, together with the unit circle for reference, makes this easy to check visually:

```{r sphere-plot, fig.width = 6, fig.height = 6}
plot(grp_circle$embedding, asp = 1, pch = 19, col = "steelblue",
     xlab = "x", ylab = "y", main = "Circular group embedding")
theta <- seq(0, 2 * pi, length.out = 200)
lines(cos(theta), sin(theta), col = "gray70")
text(grp_circle$embedding, labels = rownames(grp_circle$embedding),
     pos = 3, cex = 0.7)
```

Points should fall almost exactly on the reference circle — the fit constrains them there — but check that they aren't all bunched together, which would indicate the optimizer struggled even with the warm start (try increasing `max_epochs`, or confirm the underlying structure is actually close to circular in the first place).

### A worked example: recovering a known circular structure

To build intuition for what a good fit looks like, it helps to see the recipe recover a *known* circular arrangement from synthetic data, since with real data you don't have ground truth to check against.

```{r sphere-synthetic}
set.seed(0)
n_items <- 16
true_angle <- sort(runif(n_items, 0, 2 * pi))

# Triplet judgments generated purely from position on the circle
circ_dist <- function(i, j) {
  d <- abs(true_angle[i] - true_angle[j])
  pmin(d, 2 * pi - d)
}
n_triplets <- 4000
trip <- t(sapply(seq_len(n_triplets), function(k) {
  idx <- sample(seq_len(n_items), 3)
  h <- idx[1]; a <- idx[2]; b <- idx[3]
  if (circ_dist(h, a) < circ_dist(h, b)) c(h, a, b) - 1L else c(h, b, a) - 1L
}))
split <- floor(0.8 * n_triplets)

fit <- train_embedding(
  X_train = trip[1:split, ], X_test = trip[(split + 1):n_triplets, ],
  d = 2L, geometry = "sphere", max_epochs = 20000L, random_state = 0L
)

true_xy <- cbind(cos(true_angle), sin(true_angle))
aligned <- align.embeddings(list(true_xy, fit$embedding), scl = FALSE)
recovered_aligned <- aligned[[2]]

rmse <- sqrt(mean(rowSums((true_xy - recovered_aligned)^2)))
rmse
```

Comparing raw `atan2()` angles (or their ranks) between the true and recovered embeddings does *not* work here: the recovered circle's rotation and reflection are arbitrary, so `atan2()`'s wraparound point can fall anywhere relative to the true item ordering, and a linear/rank correlation between the two angle sequences can come out strongly negative even when the circular structure was recovered correctly. `align.embeddings()` avoids this by finding the rotation and reflection (and, if `scl = TRUE`, scale) that best superimposes the two embeddings before comparing them — the same approach the package uses to align embeddings across participants. The RMS distance after alignment is close to `0` for a good recovery; on a unit circle the largest possible distance between two points is `2`.

### Combining with `estimate_dimensionality()`

`estimate_dimensionality()` also accepts `geometry` and `radius`, so you can compare hold-out loss for spherical vs. Euclidean fits at the same set of dimensions:

```{r sphere-dimest}
dim_est_sphere <- estimate_dimensionality(
  triplet_list = icon_triplets,
  dims         = 1:6,
  n_restarts   = 5L,
  geometry     = "sphere"
)
dim_est_sphere$summary
```

Comparing this against a `geometry = "euclidean"` run at the same `dims` is a reasonable way to check whether constraining the data to a sphere costs you meaningfully in hold-out accuracy — if it doesn't, that's evidence the spherical structure is a genuine (or at least harmless) description of the data rather than a misspecification.

---

## Choosing the number of dimensions

The number of embedding dimensions `d` is a hyperparameter that must be set before fitting. Too few dimensions underfit the similarity structure; too many overfit noise. The right approach is to treat `d` as an unknown, fit embeddings at several values, and choose the smallest `d` whose hold-out loss is not meaningfully higher than the best.

Because the optimizer is stochastic, hold-out loss varies across runs at the same `d`. A single fit per dimensionality is not reliable — a good run at `d = 3` might beat a bad run at `d = 4` by chance. `estimate_dimensionality()` addresses this by fitting multiple independent random restarts at each `d` and recording the best hold-out loss from each restart.

### Running the grid

```{r dim-estimate}
dim_est <- estimate_dimensionality(
  triplet_list = icon_triplets,
  dims         = 1:8,       # dimensionalities to evaluate
  n_restarts   = 10L,       # independent random restarts per d
  max_epochs   = 50000L,
  seed         = 42L
)
```

Progress messages are printed as each restart completes. The function returns a named list with two elements.

`$results` has one row per (dimension, restart):

```{r dim-results}
head(dim_est$results)
#>   d restart      loss epoch
#> 1 1       1 0.6022...   ...
#> 2 1       2 0.6019...   ...
#> ...
```

`$summary` has one row per dimension, with the `best_d` flag identifying the recommended choice:

```{r dim-summary}
dim_est$summary
#>   d mean_loss  min_loss    sd_loss best_d
#> 1 1    0.601     0.601     0.001   FALSE
#> 2 2    0.512     0.510     0.003   FALSE
#> 3 3    0.448     0.445     0.004    TRUE   # <-- recommended
#> 4 4    0.447     0.443     0.005   FALSE
#> ...
```

### How `best_d` is determined

The recommendation uses the **one-standard-error rule**: find the dimensionality with the lowest mean hold-out loss, compute one standard error of that mean (i.e. `sd_loss / sqrt(n_restarts)`), and then flag the *smallest* `d` whose mean loss falls within that margin. This favours parsimony — it selects the simplest model that is statistically indistinguishable from the best.

For that standard error to mean anything, restarts have to disagree for a real reason — not just because of which random initialization the optimizer happened to start from. So every restart evaluates against its own freshly resampled `internal_test` subset of the training data (see `internal_test_frac` below), rather than one fixed hold-out shared by every restart. If every restart shared the same hold-out, `sd_loss` would reflect only optimization noise and understate real uncertainty, which tends to make the one-SE rule pick a higher `d` than is actually well-supported — a fit at, say, `d = 5` can always find some idiosyncrasy of one particular hold-out to exploit that `d = 3` can't. Resampling per restart gives `sd_loss` a genuine data-resampling component instead. A given restart's `internal_test` sample is identical across every `d`, so comparisons between dimensions for a matched restart stay apples-to-apples.

`internal_test_frac` (default `0.1`) controls how much of the training pool each restart holds out this way. What actually stabilizes the per-restart loss estimate is the *absolute* number of held-out triplets, not the fraction — with a training pool in the thousands, even 10% typically leaves plenty of `internal_test` triplets, so there's rarely a reason to reach for a more conventional 20% split at the cost of starving the fit itself of data. The `sampleSet == "test"` items in your data are never touched by `estimate_dimensionality()` at all — they stay reserved for evaluating the finally-selected embedding (e.g. via `run_group_embedding_from_list()`), not for choosing `d`.

### Visualizing the results

A simple plot shows mean loss by dimension with ±1 SD error bars. The vertical dashed line marks the recommended `d`.

```{r dim-plot, fig.width = 6, fig.height = 4}
s <- dim_est$summary

plot(s$d, s$mean_loss,
     type = "b", pch = 19,
     xlab = "Dimensions (d)",
     ylab = "Mean test loss",
     main = "Hold-out loss by dimensionality")

arrows(s$d, s$mean_loss - s$sd_loss,
       s$d, s$mean_loss + s$sd_loss,
       angle = 90, code = 3, length = 0.05, col = "gray50")

abline(v = s$d[s$best_d], lty = 2, col = "steelblue")
```

Look for an **elbow**: a point where increasing `d` no longer reduces loss substantially. The one-SE rule formalises this by returning the smallest `d` in the flat region of the curve.

### Group vs. individual mode

By default (`group = TRUE`), all participants' trials are pooled into a single dataset and one dimensionality search is run on the combined data. This is the right choice when you want to select a single `d` for a group embedding, and it is faster because only one set of models is trained.

```{r dim-estimate-group}
# Default: pool all participants (group = TRUE is the default)
dim_est_grp <- estimate_dimensionality(
  triplet_list = icon_triplets,
  dims         = 1:6,
  n_restarts   = 10L
)
dim_est_grp$summary
```

If participants may differ substantially in representational complexity — for example, if some participants produce clearly lower-dimensional structure than others — you can run the search separately for each participant with `group = FALSE`. In this mode the function returns a named list, one element per participant, each with the same `results` / `summary` structure.

```{r dim-estimate-individual}
dim_est_ind <- estimate_dimensionality(
  triplet_list = icon_triplets,
  dims         = 1:6,
  n_restarts   = 10L,
  group        = FALSE
)

# Result is a named list, one entry per participant
names(dim_est_ind)

# Recommended d for the first participant
dim_est_ind[[1]]$summary

# Extract the recommended d for every participant
sapply(dim_est_ind, function(x) x$summary$d[x$summary$best_d])
```

Item indices are rebuilt from each participant's own trials in individual mode, so the item space may differ across participants if not everyone saw every stimulus.

### Practical notes

The function can be slow for large datasets or many restarts because it trains `length(dims) × n_restarts` independent models (multiplied by the number of participants in individual mode). A few suggestions:

- Start with a coarse grid (e.g. `dims = c(1, 2, 3, 5, 8)`) and `n_restarts = 5` to get a rough picture, then refine around the elbow.
- Use `max_epochs` and `tol_window` values consistent with the values you plan to use for the final embedding.
- If a GPU is available, pass `device = "cuda"` (or `"mps"` on Apple Silicon) to speed up each individual fit.
- Each (d, restart) pair is independent, so the function can be parallelised — see the next section.

---

## Running in parallel

Because every (d, restart) fit is completely independent, `estimate_dimensionality()` can distribute work across multiple cores or a compute cluster with no changes to your analysis code. Parallelism is controlled via the [`future`](https://future.futureverse.org/) framework: you set a *plan* once before calling the function, and the function uses however many workers that plan provides.

Install the required packages once:

```{r parallel-install}
install.packages(c("future", "future.apply", "progressr"))
```

### Prerequisites for parallel execution

**The package must be installed, not just loaded with `load_all()`.** Each worker is a fresh R process that loads `tripletTools` as a regular package. If you are developing the package and have only run `devtools::load_all()`, the workers will fail with "there is no package called 'tripletTools'". Run `devtools::install()` first, then load the package with `library(tripletTools)` before setting up a parallel plan.

**Use `device = "cpu"` when running in parallel.** Multiple workers simultaneously competing for a single GPU causes resource contention and rarely speeds things up. Let the cores do the parallelism and leave GPU acceleration for single-model fits.

**Progress monitoring.** Because worker output is not forwarded to the main session, the `verbose` progress messages are suppressed automatically. Install `progressr` to get a progress bar showing completions instead. Wrap the call in `with_progress()` — this is more reliable in RStudio than `handlers(global = TRUE)`, which can fail if RStudio has already registered its own handlers:

```{r parallel-progress}
library(progressr)

with_progress({
  dim_est <- estimate_dimensionality(...)
})
```

### Local multicore

Use all available cores on your workstation (or a subset):

```{r parallel-local}
library(future)
library(progressr)

# Pass the plan name as a string — the bare symbol form (plan(multisession))
# may not be found if future is loaded but not attached in all environments.
plan("multisession", workers = 4)   # or: workers = parallel::detectCores() - 1

with_progress({
  dim_est <- estimate_dimensionality(
    triplet_list = icon_triplets,
    dims         = 1:8,
    n_restarts   = 10L,
    max_epochs   = 50000L,
    seed         = 42L,
    device       = "cpu"   # use CPU in parallel — GPU contention rarely helps
  )
})

plan("sequential")   # restore serial execution when done
```

With 4 workers and 80 total fits (8 dims × 10 restarts), wall time drops to roughly one quarter of the serial time.

`multisession` launches separate R processes, so it works on Windows, macOS, and Linux. If you are on Linux or macOS and prefer lower overhead, `plan("multicore")` uses forked processes instead. Note that each worker initialises Python and the conda environment on startup, so there is some fixed overhead before the first job completes and the progress bar begins updating.

### HTCondor cluster

If you have access to an HTCondor cluster (e.g. UW–Madison's CHTC), the dimensionality-search-to-final-fit pipeline above (and two other embedding-related workflows — a group-difference permutation test, and per-participant individual embeddings) can be dispatched to it directly instead of running via `future`/`plan()` on your own machine. This uses a completely separate mechanism from the local multicore approach above: a small Python driver on the submit node, with every actual fit running inside the `tripletTools` container image on Condor execute nodes — **no R installation is needed on the submit node**, only Python 3 and PyYAML.

See `vignette("condor_workflows_vignette")` for the full deployment guide: initial submit-node setup, the general procedure common to all three workflows, the container image these jobs run in, troubleshooting a job that seems stuck, and a dedicated section per workflow (including this one, `condor_workflow.py`).

### Checking the active plan

At any point you can confirm which plan is in effect:

```{r plan-check}
library(future)
plan()   # prints a description of the current plan
```

`plan(sequential)` (the default) runs everything serially and requires neither `future` nor `future.apply` — all other `tripletTools` functions are unaffected.
