---
title: "Computing Triplet Embeddings"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Computing Triplet Embeddings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Overview

This vignette explains how to compute similarity embeddings from triplet judgment data collected with this package. Embeddings are computed by a Python backend that uses PyTorch to optimize the crowd kernel (CKL) noise model. From R you interact with this backend through three functions:

| Function | What it does |
|---|---|
| `setup_python_env()` | One-time installation of the Python environment |
| `estimate_dimensionality()` | Fits embeddings across a range of dimensions to help choose `d` |
| `run_group_embedding_from_list()` | Trains a single embedding on all participants' data combined |
| `run_embeddings_from_list()` | Trains one embedding per participant plus a group embedding |

All functions accept data in the named-list format used throughout `tripletTools` (e.g. `icon_triplets`).

```{r load}
library(tripletTools)
```

---

## Step 1: One-time Python environment setup

The embedding pipeline runs inside a self-contained conda environment. You only need to do this once after installing the package. It installs PyTorch, NumPy, pandas, scikit-learn, scipy, and skorch into an environment called `"triplet-embeddings"`.

```{r setup-env}
setup_python_env()
```

PyTorch is a large download (300–800 MB depending on your platform), so the first run may take several minutes. On future R sessions you do **not** need to call `setup_python_env()` again — the environment is detected and activated automatically when you load the package.

If conda is not found on your system, `setup_python_env()` will stop with a message directing you to the Miniconda installation page.

---

## Step 2: Computing a group embedding

`run_group_embedding_from_list()` pools the judgments from all participants and trains a single embedding that captures the shared similarity structure across the group. This is the fastest option and is a natural starting point before computing individual embeddings.

```{r group-embedding}
grp <- run_group_embedding_from_list(
  triplet_list = icon_triplets,
  d            = 3L,       # number of dimensions
  max_epochs   = 50000L
)
```

The function prints training progress every 100 epochs, showing train loss, test loss, train accuracy, and test accuracy. Training stops early if the test loss has not improved meaningfully for 10,000 consecutive epochs.

### What is returned

`run_group_embedding_from_list()` returns a named list:

```{r group-result-str}
str(grp)
#> List of 3
#>  $ embedding: num [1:32, 1:3] ...
#>   ..- attr(*, "dimnames")=List of 2
#>   .. ..$ : chr [1:32] "bdbwb" "bdbwb2" ...   # item names as row names
#>   .. ..$ : chr [1:3] "dim_0" "dim_1" "dim_2"
#>  $ loss     : num 0.412
#>  $ history  :'data.frame': 847 obs. of 5 variables:
#>   ..$ epoch     : int [1:847] 0 1 2 ...
#>   ..$ train_loss: num [1:847] ...
#>   ..$ test_loss : num [1:847] ...
#>   ..$ train_acc : num [1:847] ...
#>   ..$ test_acc  : num [1:847] ...
```

- `embedding` — a matrix with one row per stimulus (item names as row names) and `d` columns (`dim_0`, `dim_1`, …).
- `loss` — the best test loss achieved during training.
- `history` — a data frame recording train/test loss and accuracy at each epoch, useful for diagnosing convergence.

### Inspecting the result

```{r group-inspect}
# First few rows of the embedding
head(grp$embedding)

# Best test loss
cat("Best test loss:", round(grp$loss, 3), "\n")

# Did training converge? Plot the test loss curve.
plot(grp$history$epoch, grp$history$test_loss,
     type = "l", col = "steelblue",
     xlab = "Epoch", ylab = "Test loss",
     main = "Group embedding – training curve")
```

### Visualizing the embedding

Once you have the group embedding, you can use `plot_pics()` to position stimulus images at their coordinates in the learned similarity space (first two dimensions shown):

```{r group-plot, fig.width = 8, fig.height = 8}
plot_pics(grp$embedding[, 1:2], icon_pics,
          psize = 0.04,
          xlab  = "Dimension 1",
          ylab  = "Dimension 2",
          main  = "Group embedding")
```

Items that appear close together were frequently judged as similar by participants overall.

The group embedding can also be used with the full set of `tripletTools` analysis functions. For example, to evaluate how well it predicts held-out judgments for participant 1:

```{r group-hoacc}
acc <- get.hoacc(grp$embedding, icon_triplets[[1]])
cat("Hold-out accuracy (participant 1):", round(acc, 3), "\n")
```

---

## Step 3: Computing individual embeddings

`run_embeddings_from_list()` trains a separate embedding for each participant and then trains an additional group embedding on all participants' data combined.  This takes longer but lets you study individual differences in how participants represent the stimulus set.

```{r ind-embeddings}
results <- run_embeddings_from_list(
  triplet_list = icon_triplets,
  output_dir   = "embeddings_output",   # CSVs are also saved here
  d            = 3L,
  max_epochs   = 50000L
)
```

Output CSV files are written to `output_dir` as a side-effect:

| File | Contents |
|---|---|
| `embeddings.csv` | All per-participant and group embeddings concatenated |
| `embeddings_group.csv` | Group embedding only |
| `model_history.csv` | Training diagnostics for each participant and the group |

### What is returned

`run_embeddings_from_list()` returns a named list:

```{r ind-result-str}
str(results, max.level = 2)
#> List of 3
#>  $ individual:List of 6
#>   ..$ 3n7ggxph: num [1:32, 1:3] ...   # one matrix per participant
#>   ..$ b5wma4no: num [1:32, 1:3] ...
#>   ..$ d8mmm1qn: num [1:32, 1:3] ...
#>   ..$ jn7bbjc0: num [1:32, 1:3] ...
#>   ..$ pbby694o: num [1:32, 1:3] ...
#>   ..$ sc2xbd6w: num [1:32, 1:3] ...
#>  $ group     : num [1:32, 1:3] ...    # group embedding matrix
#>  $ history   :'data.frame': 7 obs. of 6 variables:
#>   ..$ worker_id              : chr [1:7] "3n7ggxph" ... "group"
#>   ..$ lowest_loss            : num [1:7] ...
#>   ..$ epoch                  : int [1:7] ...
#>   ..$ counter_from_last_update: int [1:7] ...
#>   ..$ n_train_triplets       : int [1:7] ...
#>   ..$ n_test_triplets        : int [1:7] ...
```

- `individual` — a named list of matrices, one per participant, in the same format as `grp$embedding` above.  Element names match the `worker_id` labels in your data.
- `group` — a single matrix containing the group embedding.
- `history` — a data frame with one row per participant plus one row for the group, summarising training diagnostics.

### Inspecting training quality

The `history` data frame gives a quick overview of how well each participant's model converged:

```{r history-inspect}
results$history[, c("worker_id", "lowest_loss", "epoch", "n_train_triplets")]
```

Participants who stopped early (small `epoch`) and achieved a low `lowest_loss` are well-modelled. If a participant's `lowest_loss` is much higher than the others, their data may be noisier or they may have fewer trials.

### Working with individual embeddings

The individual embeddings are stored in `results$individual` and have the same structure as the group embedding: a matrix with item row names and columns `dim_0`, `dim_1`, ….  This means they are compatible with all existing `tripletTools` functions.

```{r ind-use}
# Hold-out prediction accuracy for each participant using their own embedding
sapply(names(results$individual), function(wid) {
  get.hoacc(results$individual[[wid]], icon_triplets[[wid]])
})

# Prediction matrix: how well does each participant's embedding predict
# every other participant's held-out judgments?
pmat <- get.prediction.matrix(results$individual, icon_triplets)

# Do participants predict their own data better than others predict it?
cat("Mean self-prediction accuracy:  ", round(mean(diag(pmat)), 3), "\n")
cat("Mean other-prediction accuracy: ", round(mean(pmat[row(pmat) != col(pmat)]), 3), "\n")
```

### Visualizing individual embeddings

```{r ind-plot, fig.width = 8, fig.height = 8}
emb1 <- results$individual[[1]]

plot_pics(emb1[, 1:2], icon_pics,
          psize = 0.04,
          xlab  = "Dimension 1",
          ylab  = "Dimension 2",
          main  = paste("Individual embedding –", names(results$individual)[1]))
```

---

## Choosing between the two functions

| | `run_group_embedding_from_list()` | `run_embeddings_from_list()` |
|---|---|---|
| **Speed** | Fast — one model | Slow — one model per participant + group |
| **Output** | Single group embedding | Per-participant + group embeddings |
| **Best for** | Quick summary of shared structure; pilot analyses | Studying individual differences; full analysis |
| **Saved files** | None | Three CSVs in `output_dir` |

If you only need the group embedding, prefer `run_group_embedding_from_list()`. If you want to study whether participants differ in their representations — using tools like `get.prediction.matrix()`, `get.rep.dist()`, or `pacc.by.cluster()` — run `run_embeddings_from_list()` to obtain individual embeddings as well.

---

## Controlling training

Both functions accept the same set of hyperparameters:

| Parameter | Default | Meaning |
|---|---|---|
| `d` | `5` | Number of embedding dimensions |
| `max_epochs` | `50000` | Maximum training epochs |
| `tolerance` | `1e-4` | Minimum loss improvement to reset the early-stopping counter |
| `tol_window` | `10000` | Epochs without improvement before stopping |
| `seed` | `222` | Random seed for reproducibility |

For exploratory analyses, the defaults are a reasonable starting point. For a final analysis you may want to increase `max_epochs` and `tol_window` to allow longer convergence, especially with large datasets or high `d`.

To inspect the training curve after fitting:

```{r training-curve}
plot(grp$history$epoch, grp$history$test_loss,
     type = "l", col = "steelblue", lwd = 2,
     xlab = "Epoch", ylab = "Test loss",
     main = "Convergence check")
lines(grp$history$epoch, grp$history$train_loss,
      col = "tomato", lwd = 2, lty = 2)
legend("topright",
       legend = c("Test", "Train"),
       col    = c("steelblue", "tomato"),
       lty    = c(1, 2), lwd = 2)
```

A well-converged model shows test loss flattening well before `max_epochs`. If the loss is still falling at the end, increase `max_epochs`.

---

## Choosing the number of dimensions

The number of embedding dimensions `d` is a hyperparameter that must be set before fitting. Too few dimensions underfit the similarity structure; too many overfit noise. The right approach is to treat `d` as an unknown, fit embeddings at several values, and choose the smallest `d` whose hold-out loss is not meaningfully higher than the best.

Because the optimizer is stochastic, hold-out loss varies across runs at the same `d`. A single fit per dimensionality is not reliable — a good run at `d = 3` might beat a bad run at `d = 4` by chance. `estimate_dimensionality()` addresses this by fitting multiple independent random restarts at each `d` and recording the best hold-out loss from each restart.

### Running the grid

```{r dim-estimate}
dim_est <- estimate_dimensionality(
  triplet_list = icon_triplets,
  dims         = 1:8,       # dimensionalities to evaluate
  n_restarts   = 10L,       # independent random restarts per d
  max_epochs   = 50000L,
  seed         = 42L
)
```

Progress messages are printed as each restart completes. The function returns a named list with two elements.

`$results` has one row per (dimension, restart):

```{r dim-results}
head(dim_est$results)
#>   d restart      loss epoch
#> 1 1       1 0.6022...   ...
#> 2 1       2 0.6019...   ...
#> ...
```

`$summary` has one row per dimension, with the `best_d` flag identifying the recommended choice:

```{r dim-summary}
dim_est$summary
#>   d mean_loss  min_loss    sd_loss best_d
#> 1 1    0.601     0.601     0.001   FALSE
#> 2 2    0.512     0.510     0.003   FALSE
#> 3 3    0.448     0.445     0.004    TRUE   # <-- recommended
#> 4 4    0.447     0.443     0.005   FALSE
#> ...
```

### How `best_d` is determined

The recommendation uses the **one-standard-error rule**: find the dimensionality with the lowest mean hold-out loss, compute one standard error of that mean (i.e. `sd_loss / sqrt(n_restarts)`), and then flag the *smallest* `d` whose mean loss falls within that margin. This favours parsimony — it selects the simplest model that is statistically indistinguishable from the best.

### Visualizing the results

A simple plot shows mean loss by dimension with ±1 SD error bars. The vertical dashed line marks the recommended `d`.

```{r dim-plot, fig.width = 6, fig.height = 4}
s <- dim_est$summary

plot(s$d, s$mean_loss,
     type = "b", pch = 19,
     xlab = "Dimensions (d)",
     ylab = "Mean test loss",
     main = "Hold-out loss by dimensionality")

arrows(s$d, s$mean_loss - s$sd_loss,
       s$d, s$mean_loss + s$sd_loss,
       angle = 90, code = 3, length = 0.05, col = "gray50")

abline(v = s$d[s$best_d], lty = 2, col = "steelblue")
```

Look for an **elbow**: a point where increasing `d` no longer reduces loss substantially. The one-SE rule formalises this by returning the smallest `d` in the flat region of the curve.

### Practical notes

The function can be slow for large datasets or many restarts because it trains `length(dims) × n_restarts` independent models. A few suggestions:

- Start with a coarse grid (e.g. `dims = c(1, 2, 3, 5, 8)`) and `n_restarts = 5` to get a rough picture, then refine around the elbow.
- Use `max_epochs` and `tol_window` values consistent with the values you plan to use for the final embedding.
- If a GPU is available, pass `device = "cuda"` (or `"mps"` on Apple Silicon) to speed up each individual fit.
- `estimate_dimensionality()` trains only a **group** embedding (pooling all participants) at each `d`. This is usually the right choice for selecting `d`, and it is faster than fitting per-participant models.
- Each (d, restart) pair is independent, so the function can be parallelised — see the next section.

---

## Running in parallel

Because every (d, restart) fit is completely independent, `estimate_dimensionality()` can distribute work across multiple cores or a compute cluster with no changes to your analysis code. Parallelism is controlled via the [`future`](https://future.futureverse.org/) framework: you set a *plan* once before calling the function, and the function uses however many workers that plan provides.

Install the required packages once:

```{r parallel-install}
install.packages(c("future", "future.apply"))
```

### Local multicore

Use all available cores on your workstation (or a subset):

```{r parallel-local}
library(future)
plan(multisession, workers = 4)   # or: workers = parallel::detectCores() - 1

dim_est <- estimate_dimensionality(
  triplet_list = icon_triplets,
  dims         = 1:8,
  n_restarts   = 10L,
  max_epochs   = 50000L,
  seed         = 42L
)

plan(sequential)   # restore serial execution when done
```

With 4 workers and 80 total fits (8 dims × 10 restarts), wall time drops to roughly one quarter of the serial time.

`multisession` launches separate R processes, so it works on Windows, macOS, and Linux.  If you are on Linux or macOS and prefer lower overhead, `plan(multicore)` uses forked processes instead.

### HTCondor cluster

If you have access to an HTCondor cluster (e.g. UW–Madison's CHTC), install the `future.batchtools` package, which bridges `future` to HTCondor's job scheduler:

```{r condor-install}
install.packages("future.batchtools")
```

You also need a one-time cluster template file that tells `future.batchtools` how to construct HTCondor submit files for your site.  Save the following as `~/.batchtools.condor.tmpl` (adjust `request_memory` and any site-specific lines for your cluster):

```
universe     = vanilla
executable   = `r shQuote(file.path(R.home("bin"), "Rscript"))`
arguments    = -e 'batchtools::doJobCollection("$(job.collection)")'

transfer_input_files  = $(job.collection)
should_transfer_files = YES
when_to_transfer_output = ON_EXIT

request_cpus   = 1
request_memory = 4GB
request_disk   = 2GB

log    = $(log)
output = $(output)
error  = $(error)

queue
```

Then submit your dimensionality search exactly as before, just with a different plan:

```{r condor-run}
library(future.batchtools)

plan(batchtools_condor, workers = 80)   # up to 80 simultaneous jobs

dim_est <- estimate_dimensionality(
  triplet_list = icon_triplets,
  dims         = 1:8,
  n_restarts   = 10L,
  max_epochs   = 50000L,
  seed         = 42L
)

plan(sequential)
```

`future.batchtools` handles job submission, monitors completion, retrieves results, and assembles them — you get back the same `dim_est` list as in the serial case.

**Prerequisite:** every worker node must have R, `tripletTools`, and the `triplet-embeddings` conda environment already installed.  This is a one-time setup task; ask your cluster administrator or follow your site's software installation guide.

### Checking the active plan

At any point you can confirm which plan is in effect:

```{r plan-check}
library(future)
plan()   # prints a description of the current plan
```

`plan(sequential)` (the default) runs everything serially and requires neither `future` nor `future.apply` — all other `tripletTools` functions are unaffected.
