---
title: "tripletTools Overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tripletTools Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 6
)
```

## Introduction

This package is designed to help analyze and visualize results from *triadic comparisons* (or *triplet tasks*), which are used to characterize the structure of mental representations. In a triplet task, a participant sees a *referent* item and two *option* items, and must decide which option is most similar to the referent. From many such judgments across many items and participants, one can compute a *similarity embedding*: a low-dimensional coordinate space in which items that are frequently judged as similar are placed nearby.

`tripletTools` does not collect data or compute embeddings — those steps occur outside R. Instead it provides tools for:

- Loading triplet and embedding data files
- Assessing participant data quality
- Measuring inter-subject agreement on shared trials
- Visualizing embeddings
- Evaluating how well an embedding predicts held-out judgments
- Quantifying individual differences in representation
- Clustering participants by representational similarity

This vignette illustrates each of these steps using the example dataset included with the package.

```{r load}
library(tripletTools)
```

---

## The example dataset

The package includes objects from a triplet study on 32 icon images of faces and buildings. The icons varied in category (face/building), time of day (day/night), and additional features (gender and race for faces; size and kind for buildings). Six participants each judged approximately 230 triplets online.

| Object | Description |
|---|---|
| `icon_triplets` | Named list of 6 data frames, one per participant, containing trial-by-trial judgments |
| `icon_emb_ind` | Named list of 6 matrices, one per participant, containing 3-D embedding coordinates |
| `icon_emb_group` | Single data frame containing a 3-D group embedding |
| `icon_pics` | Named list of 32 PNG images, one per stimulus icon |

### Triplet data

Each element of `icon_triplets` is one participant's trial data:

```{r triplets-head}
head(icon_triplets[[1]])
```


The key columns are `Center` (the referent item), `Left` and `Right` (the two options), `Answer` (the participant's choice), `sampleAlg` (how the trial was sampled: `random`, `check`, or `validation`), and `sampleSet` (whether the trial was used to fit the embedding — `train` — or held out for testing — `test`).

### Embedding data

Each element of `icon_emb_ind` is a 32-row matrix of 3-D embedding coordinates, with row names equal to stimulus names:

```{r embeddings-head}
head(icon_emb_ind[[1]])
```

---

## Data quality assessment

Before analyzing results it is good practice to check whether participants engaged with the task. `get.participant.summary` produces a per-participant summary including the number of trials completed, mean accuracy on *check trials* (easy triplets with an obvious answer used as an attention check), and mean log response time.

```{r quality}
psummary <- get.participant.summary(icon_triplets, mintrial = 230)
head(psummary)
```

The output includes a `keep` column flagging participants who fall below the specified thresholds (the specified minimum number of trials, check-trial accuracy ≥ 0.80, mean log RT > 0). Participants failing these criteria should be reviewed before including their data in downstream analyses.

---

## Inter-subject agreement on validation trials

A subset of trials — the *validation* trials — are drawn from a fixed pool shared across all participants, meaning the same triplets appear in multiple participants' datasets. This allows us to measure inter-subject consistency: for each validation triplet we identify the *majority vote* (the option chosen by most participants who saw it) and the proportion of participants who agreed with that majority.

`make.vmat` computes this from the full triplet list:

```{r vmat}
vmat <- make.vmat(icon_triplets)

# One row per unique validation triplet
head(vmat$majority)
```

The `pmaj` column shows the proportion of participants who agreed with the majority choice. Values near 1.0 indicate near-universal agreement; values near 0.5 indicate a closely split decision.

```{r vmat-hist, fig.cap="Distribution of inter-subject agreement on validation trials."}
hist(vmat$majority$pmaj,
     breaks = 10,
     main   = "Inter-subject agreement on validation trials",
     xlab   = "Proportion agreeing with majority vote",
     col    = "steelblue", border = "white")
abline(v = 0.5, lty = 2, col = "red")
```

The `bysbj` element is a participant × triplet matrix recording whether each participant agreed with the majority vote (1), disagreed (0), or did not see the triplet (NA). The mean across triplets gives each participant's overall agreement rate:

```{r vmat-mean}
barplot(rowMeans(vmat$bysbj, na.rm = TRUE), 
        beside = T, 
        ylim = c(0, 1.0),
        xlab = "Participant",
        ylab = "Agreement",
        cex.names = 0.5,
        las=2)
box()
abline(h = 0.5, lty = 2)
```

Participants can vary quite a bit as to how well they agree with the majority vote, indicating potential individual differences in how people view these stimuli.

---

## Visualizing a 2-D embedding

`plot_pics` plots images as points in a scatterplot, positioned at their embedding coordinates. This gives an immediate visual impression of how the faces are organized in the learned similarity space.

```{r plot-emb, fig.cap="3-D embedding for participant 1, with icon images as points (first two dimensions shown).", fig.width = 8, fig.height = 8}
emb1 <- icon_emb_ind[[1]]

plot_pics(emb1[,1:2], icon_pics,
          psize = 0.04,
          xlab  = "Dimension 1",
          ylab  = "Dimension 2",
          main  = "Embedding – participant 1")
```

Icons that appear close together were frequently judged as similar to one another by this participant.

---

## Evaluating embedding quality: hold-out prediction accuracy

After computing an embedding we want to know how well it captures the participant's actual judgments. For each held-out (*test*) triplet we can ask: does the embedding correctly predict which option the participant chose? The predicted choice is whichever option is closer to the referent in the embedding space.

`get.hoacc` returns the proportion of held-out trials for which the embedding's prediction matches the participant's answer:

```{r hoacc}
acc1 <- get.hoacc(icon_emb_ind[[1]], icon_triplets[[1]])
cat("Hold-out prediction accuracy (participant 1):", round(acc1, 3), "\n")
```

Chance performance is 0.50. Values above approximately 0.60 are generally considered reasonable for a 2-D embedding.

`test.model` returns trial-level predictions, making it easy to inspect specific errors or compute accuracy on any subset of trials:

```{r test-model}
result     <- test.model(icon_emb_ind[[1]], icon_triplets[[1]])
test_rows  <- result$sampleSet == "test"
mean(result$ModPred[test_rows] == result$Answer[test_rows], na.rm=TRUE)
```

### Prediction strength

`model.strength` computes, for each triplet, how *decisively* the embedding favors one option over the other. The metric is max(d1, d2) / (d1 + d2), where d1 and d2 are the distances from the referent to each option. It ranges from 0.5 (options equidistant from the referent) to 1.0 (one option far closer than the other).

If the embedding is well-calibrated, trials where it is most confident should also be the trials where participants agree most strongly. We can test this using the validation trials, for which we have an inter-subject agreement measure. *NOTE*: typically one would evaluate this for a group embedding, since validation statistics are computed across the whole group. Here we illustrate using the individual embedding from Participant 1; see `icon_emb_group` for the group embedding of these stimuli:

```{r model-strength, fig.cap="Embedding prediction strength vs. inter-subject agreement on validation trials."}
# Validation trials for participant 1
vtrips <- subset(icon_triplets[[1]], sampleAlg == "validation")

# Prediction strength from participant 1's embedding
pstrength <- model.strength(icon_emb_ind[[1]], vtrips)

# Locate matching inter-subject agreement values
tnames  <- make.tripnames(vtrips)
maj_idx <- match(tnames, vmat$majority$triplet)
pmaj    <- vmat$majority$pmaj[maj_idx]

# Plot the relationship
plot(pstrength, pmaj,
     xlab = "Embedding prediction strength",
     ylab = "Inter-subject agreement",
     pch  = 16, col = rgb(0, 0, 0.8, 0.4),
     main = "Strength vs. inter-subject agreement")
abline(lm(pmaj ~ pstrength), col = "red")

text(.5, .96, paste("r =", round(cor(pstrength, pmaj, use = "complete.obs"), 3)), adj = 0)
```

A positive correlation indicates that the embedding correctly identifies which triplets have clearer, more consistent answers. 

---

## Individual differences: the prediction matrix

A central question in many triplet studies is whether participants differ reliably in their representations. The *prediction matrix* approach addresses this: we use each participant's embedding to predict every other participant's held-out judgments. If individual differences are real and systematic, a participant's *own* embedding should predict their judgments better than another participant's embedding does.

`get.prediction.matrix` computes this for all pairs:

```{r predmat}
pmat <- get.prediction.matrix(icon_emb_ind, icon_triplets)
```

The result is a 6 × 6 matrix. Entry [i, j] is the accuracy with which participant i's embedding predicts participant j's held-out judgments. The diagonal contains each participant's *self-prediction* accuracy.

```{r predmat-compare}
cat("Mean self-prediction accuracy:  ", round(mean(diag(pmat)), 3), "\n")
cat("Mean other-prediction accuracy: ", round(mean(pmat[row(pmat) != col(pmat)]), 3), "\n")
```

`z.pred.mat` converts each diagonal entry to a z-score relative to the other entries in its row, expressing how much better (or worse) a participant's own embedding predicts their data compared to other participants' embeddings:

```{r zpredmat}
zscores <- z.pred.mat(pmat)

cat("Mean z-score of self-prediction:", round(mean(zscores, na.rm = TRUE), 3), "\n")
cat("Proportion with positive z-score:", round(mean(zscores > 0, na.rm = TRUE), 3), "\n")
```

If z-scores are reliably positive across participants, this is evidence of meaningful individual differences in the representation of the stimulus set.

---

## Representational distances between participants

`get.rep.dist` computes the pairwise *procrustes distance* between all pairs of embeddings. Two embeddings that can be brought into near-perfect alignment by rotation, scaling, and reflection have a low distance; embeddings that remain dissimilar after alignment have a high distance.

```{r repdist}
repdist <- get.rep.dist(icon_emb_ind)
```

We can cluster participants by their representational distances using standard hierarchical clustering:

```{r hclust, fig.cap="Participants clustered by pairwise representational distance."}
#Compute hierarchical cluster tree:
hc     <- hclust(as.dist(repdist), method = "ward.D") 

#Cut tree into 2 clusters:
clusts <- cutree(hc, k = 2)

#Plot tree and highlight groups created by cutting:
plot(hc,
     labels = FALSE,
     main   = "Participant clustering by representational distance",
     xlab   = "", sub = "")
rect.hclust(hc, k = 2, border = c("tomato", "steelblue"))
```

### Mean embedding per cluster

`get.group.list.mean` aligns all embeddings within each cluster and returns a mean embedding for each group:

```{r group-mean}
mn_by_clust <- get.group.list.mean(icon_emb_ind, clusts)
```

Plotting each cluster's mean embedding shows whether different groups of participants organized the icons in qualitatively different ways:

```{r plot-cluster-1, fig.cap="Mean embedding for cluster 1 (first two dimensions).", fig.height= 8, fig.width = 8}

dmat <- mn_by_clust[[1]] #Copy data matrix
dmat <- dmat / max(abs(dmat)) #Max scaling

tripletTools::plot_pics(dmat, icon_pics, psize = 0.04,
          xlab = "Dimension 1", ylab = "Dimension 2",
          main = "Cluster 1 – mean embedding")
```

```{r plot-cluster-2, fig.cap="Mean embedding for cluster 2 (first two dimensions).", fig.width = 8, fig.height = 8}
dmat <- mn_by_clust[[2]] #Copy data matrix
dmat <- dmat / max(abs(dmat)) #Max scaling

tripletTools::plot_pics(dmat, icon_pics, psize = 0.04,
          xlab = "Dimension 1", ylab = "Dimension 2",
          main = "Cluster 2 – mean embedding")
```

---

## Prediction accuracy by cluster

If the clustering captures genuine individual differences, a participant's held-out judgments should be better predicted by embeddings from within their cluster than by embeddings from other clusters. `pacc.by.cluster` tests this directly using the prediction matrix:

```{r pbc}
pbc <- pacc.by.cluster(pmat, clusts, samediff = TRUE)
head(pbc)
```

Each row is one participant. The three columns give prediction accuracy from (1) their own embedding, (2) the mean of their cluster-mates' embeddings, and (3) the mean of participants outside their cluster.

```{r pbc-means}
round(colMeans(pbc), 3)
```

We can show the mean and 95% confidence interval of these values across participants as a ribbon plot:

```{r pbc-plot, fig.cap="Hold-out prediction accuracy: self, same cluster, and other cluster."}
plot_cis(pbc,
         xvals  = 1:3,
         main   = "Prediction accuracy by cluster membership",
         ylab   = "Hold-out prediction accuracy",
         xaxt   = "n",
         ylim   = c(0.5, 1.0))
axis(1, at = 1:3, labels = c("Self", "Same cluster", "Other cluster"))
abline(h = 0.5, lty = 2, col = "grey50")
```

A higher mean for "same cluster" than "other cluster" indicates that cluster membership captures something real about how participants differ in their representations.

---

## Tree visualization with images as leaves

For a richer view of the similarity structure in one participant's embedding, a hierarchical cluster tree with face images at the leaves can be more informative than a 2-D scatterplot — especially for higher-dimensional embeddings. `get.tip.coords` retrieves the tip coordinates of a phylogram plot, allowing `plot_pics` to place images at the correct positions. In this case we will visualize the hierarchical cluster plot of the single embedding computed from the full group of participants (icon_emb_group).

```{r tree-plot, fig.width = 8, fig.height = 8, fig.cap="Similarity tree for participant 1, with icon images at the leaves."}
hc_emb <- hclust(dist(icon_emb_group[,1:3]), method = "ward.D2")
pt     <- ape::as.phylo(hc_emb)

plot(pt, type = "fan", show.tip.label = FALSE,
     main = "Icon similarity tree – group")

tip_coords <- get.tip.coords()
plot_pics(tip_coords, icon_pics, newplot = FALSE, psize = 0.04)
```

Icons on nearby branches were judged as similar overall by the group.

---

## Summary

The table below maps common analysis questions to the corresponding `tripletTools` functions:

| Question | Function(s) |
|---|---|
| Did participants engage with the task? | `get.participant.summary` |
| How consistent are judgments across participants? | `make.vmat` |
| What does a participant's embedding look like? | `plot_pics` |
| How well does the embedding predict held-out judgments? | `get.hoacc`, `test.model` |
| Does the embedding correctly rank triplet difficulty? | `model.strength` |
| Are individual differences in representation reliable? | `get.prediction.matrix`, `z.pred.mat` |
| How similar are two participants' representations? | `get.rep.dist` |
| Are there subgroups with similar representations? | `get.rep.dist`, `get.group.list.mean` |
| Do cluster-mates predict each other's data better? | `pacc.by.cluster`, `plot_cis` |
| How to show a full similarity tree with images? | `get.tip.coords`, `plot_pics` |
