---
title: "Reading Triplet Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading Triplet Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tripletTools)
```

This package contains a set of functions to aid analysis of data from _triadic comparisons_ or _triplet_ tasks. See the _tripletTools Overview_ vignette for demonstrations of the various tools.

This vignette describes how to get triplet data into R in a format that works with the package. There are three common starting points:

1. **Raw jsPsych exports** — you just finished data collection and have one CSV per participant straight out of jsPsych. Use `read_raw_data()`.
2. **A legacy dataset** — you have a single CSV from an older experiment or a different collection platform, with column names that may not match the package's conventions. Use `read_legacy()`.
3. **Already-processed data** — your data are already in the standard format (e.g. previously exported by one of the functions above). Jump straight to [Data file structure and naming conventions](#data-file-structure-and-naming-conventions) below to see how to load it.

---

## Reading raw jsPsych data with `read_raw_data()`

If you collected your data with a jsPsych triplet experiment, `read_raw_data()` handles the full cleaning pipeline in one call. Point it at the directory containing the raw participant CSV files and it will:

1. Read and combine all CSVs in the directory.
2. Filter rows to experiment trials (`trial_category` equal to `"random"`, `"check"`, or `"validation"`).
3. Apply quality-control exclusions: remove participants with too few trials (`filter_incomplete`), unusually fast mean reaction times (`filter_fast_responders`), or too many failed catch trials (`filter_failed_catch`).
4. Strip file paths and extensions from stimulus names.
5. Split the comma-separated `choices` column into labelled `Left` / `Right` columns and derive `winner` / `loser`.
6. Assign a random 80/20 train/test split (configurable) via `assign_sample_sets`.
7. Optionally write a cleaned trial CSV and a stimulus-level integer-mapping CSV to disk.

```{r read-raw, eval = FALSE}
result <- read_raw_data(
  data_dir       = "experiment/raw_data/",   # folder of per-participant CSVs
  output_df      = "data/triplets_clean.csv",
  output_levels  = "data/stimulus_levels.csv",
  min_trials     = 200,       # drop participants with fewer trials than this
  min_mean_rt_ms = 200,       # drop participants with mean RT below this (ms)
  max_prop_wrong = 0.20,      # drop participants who fail >20% of catch trials
  test_prop      = 0.20,      # fraction of trials assigned to the test set
  seed           = 42
)
```

The function returns a list invisibly with two elements:

- `result$trials` — a data frame of cleaned trial-level data in the standard format (see [Data file structure](#data-file-structure-and-naming-conventions) below), ready to pass to `get.combined()` or the embedding functions.
- `result$levels` — a data frame mapping each stimulus label to its integer index and file path.

All QC thresholds default to permissive values (`min_trials = 0`, `min_mean_rt_ms = 0`, `max_prop_wrong = 1`), so calling `read_raw_data()` with only `data_dir` set will read and clean the files without excluding anyone.

---

## Reading legacy data with `read_legacy()`

If your data come from an older experiment or a different collection platform, the column names may not match the package's conventions. `read_legacy()` handles a wide range of input formats by matching column names case-insensitively and ignoring punctuation, so it can usually find the right columns even when they are named differently.

Pass the path to a single combined CSV file:

```{r read-legacy, eval = FALSE}
result <- read_legacy("data/old_experiment.csv")
```

The function recognises many common alternative column names automatically. For example:

| Standard name | Also recognised as |
|---|---|
| `worker_id` (participant) | `sessionID`, `puid`, `Participant.ID`, `sub_id`, `pid` |
| `Center` (target item) | `Target` |
| `Left` / `Right` (options) | `Option1` / `Option2` |
| `sampleAlg` | `AlgSample` |
| `sampleSet` | `Alg.Label`, `TrnTest` |

When a column cannot be found, it is derived where possible: `winner` and `loser` are inferred from `Answer`, `Left`, and `Right`; if no participant identifier is found, participants are assigned sequential IDs (`"P1"`, `"P2"`, …); if no `sampleSet` column exists, 10% of trials are randomly assigned to `"test"` and the rest to `"train"`.

`read_legacy()` always writes two output files alongside the input:

- `<basename>_v2025.csv` — standardised trial data in the package's column format.
- `<basename>_2025_levels.csv` — stimulus-level integer mapping.

The function returns a list invisibly with the paths to these two files:

```{r read-legacy-result, eval = FALSE}
result$data_file    # path to the standardised CSV
result$levels_file  # path to the levels mapping CSV
```

Once you have a standardised CSV from either function, load it into R with `get.combined()` as described in the next section.

---

## Data file structure and naming conventions

The functions make use of two kinds of data files: _triplet data files_, which contain information about each trial of a tradic comparison task, and _embedding files_, which contain the coordinates of each item in an embedding computed from triplet judgment data.

### Triplet data files

Triplet data files are `.csv` files generated by the software used to collect triplet judgment data. The first row should be a header specifying column names. Each subsequent row then records key information for each trial of a triplet experiment. Typically data from all participants in a given study are included in a single triplet data file. 

The triplet data file must be a `.csv` file and must contain columns with the following names:

- _worker\_id_: Arbitrary identifier for each participant
- _rt_: Response time for the trial
- _Center_, _Left_, _Right_: Strings indicating the items appearing in the center (target item), left side (option 1) and right side (option 2).
- _Answer_: String indicating which option the participant chose.
- _sampleAlg_: ALgorithm used to sample the item: either random, validation, or check.
- _sampleSet_: Indicates whether the triplet was used to fit the embedding (train) or not (test).

The data file can also contain any other fields. Often data will include an integer encoing of the triplet information with the following column names:

- _head_, _winner_, _loser_: Integer indices for each item appearing in the triplet

Data in this format can be read into the current session using the function `get_combined(fname)` where `fname` is the path to the data file. This function returns a named list, where each element includes the triplet judgment data from a single subject, and elements are named by the subject identified. This package includes an example dataset in this format, `icon_triplets`:

```{r}
head(icon_triplets[[1]])
```
Here you can see the triplet judgment data for each trial for the first participant in the experiment. Participants viewed the trials in the same order they are listed in the matrix.

The data from each participant is a separate element in the list, and the elements are labeled by the `worker\_id` label in the raw data file. You can see all the subject labels as follows:

```{r}
names(icon_triplets)
```

To learn more about this dataset, try `help(icon_triplets)`.

### Embedding data

Embedding data are `.csv` files containing embedding coordinates for each stimulus item in the study. Depending on the study, there may be a single _group embedding_ computed from a group of participants, or _individual embeddings_ computed separately for each participant, or both.

In both cases the `.csv` file must contain columns with the following labels:

- _item_: A string indicating the label for the item.
- _dim\_0 - dim\_k_: One column for each dimension of the embedding, numbered beginning with zero, containing a numeric value that indicates the item's location on the corresponding dimension of the embedding.

If a separate embedding was computed for each participant, then all embeddings should appear within the same `.csv` file, and this should also include the following column:

- _worker\_id_: The random participant identifier, which should be the same as the identifier used in the triplet dataset.

To read in a group-based single embedding file you can just use standard R:

```
grpemb <- read.csv("filename.csv", row.names = "item", header = T)
```

For studies with separate embeddings computed for each participant, use the `get.combined` function to read the data, setting the `eflag` flag to `TRUE` to indicate these are embeddings:

```
indemb <- get.combined("filename.csv", eflag = TRUE)
```

As with triplet data, this will create a named list where each element contains the embedding information computed for one participant. The elements are labeled by the participant id (`worker_id`). The `icon_emb_ind` object contains a list of the kind returned by this function:

```{r}
head(icon_emb_ind[[1]])
```
The row names indicate the stimulus identity and the entries indicate the coordinates of the stimulus along the first (dim_0), second (dim_1), and third (dim_2) dimensions of the embedding.
