---
title: "Reading Triplet Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading Triplet Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tripletTools)
```

This package contains a set of functions to aid analysis of data from _triadic comparisons_ or _triplet_ tasks. See the _tripletTools Overview_ vignette for demonstrations of the various tools.

This vignette describes how to read new data into R in a format that works with the package, and describes how the raw data file should be structured.

## Data file structure and naming conventions

The functions make use of two kinds of data files: _triplet data files_, which contain information about each trial of a tradic comparison task, and _embedding files_, which contain the coordinates of each item in an embedding computed from triplet judgment data.

### Triplet data files

Triplet data files are `.csv` files generated by the software used to collect triplet judgment data. The first row should be a header specifying column names. Each subsequent row then records key information for each trial of a triplet experiment. Typically data from all participants in a given study are included in a single triplet data file. 

The triplet data file must be a `.csv` file and must contain columns with the following names:

- _worker\_id_: Arbitrary identifier for each participant
- _rt_: Response time for the trial
- _Center_, _Left_, _Right_: Strings indicating the items appearing in the center (target item), left side (option 1) and right side (option 2).
- _Answer_: String indicating which option the participant chose.
- _sampleAlg_: ALgorithm used to sample the item: either random, validation, or check.
- _sampleSet_: Indicates whether the triplet was used to fit the embedding (train) or not (test).

The data file can also contain any other fields. Often data will include an integer encoing of the triplet information with the following column names:

- _head_, _winner_, _loser_: Integer indices for each item appearing in the triplet

Data in this format can be read into the current session using the function `get_combined(fname)` where `fname` is the path to the data file. This function returns a named list, where each element includes the triplet judgment data from a single subject, and elements are named by the subject identified. This package includes an example dataset in this format, `cfd_triplets`:

```{r}
head(cfd_triplets[[1]])
```
Here you can see the triplet judgment data for each trial for the first participant in the experiment. Participants viewed the trials in the same order they are listed in the matrix. 

The data from each participant is a separate element in the list, and the elements are labeled by the `worker\_id` label in the raw data file. You can see all the subject labels as follows:

```{r}
names(cfd_triplets)
```

To learn more about this dataset, try `help(cfd_triplets)`.

### Embedding data

Embedding data are `.csv` files containing embedding coordinates for each stimulus item in the study. Depending on the study, there may be a single _group embedding_ computed from a group of participants, or _individual embeddings_ computed separately for each participant.

In both cases the `.csv` file must contain columns with the following labels:

- _item_: A string indicating the label for the item.
- _dim\_0 - dim\_k_: One column for each dimension of the embedding, numbered beginning with zero, containing a numeric value that indicates the item's location on the corresponding dimension of the embedding.

If a separate embedding was computed for each participant, then all embeddings should appear within the same `.csv` file, and this should also include the following column:

- _worker\_id_: The random participant identifier, which should be the same as the identifier used in the triplet dataset.

To read in a group-based single embedding file you can just use standard R:

```
grpemb <- read.csv("filename.csv", row.names = "item", header = T)
```

For studies with separate embeddings computed for each participant, use the `get.combined` function to read the data, setting the `eflag` flag to `TRUE`:

```
indemb <- get.combined("filename.csv", eflag = TRUE)
```

As with triplet data, this will create a named list where each element contains the embedding information computed for one participant. The elements are labeled by the participant id (`worker_id`). The `cfd_embeddings` object contains a list of the kind returned by this function:

```{r}
head(cfd_embeddings[[1]])
```
The row names indicate the stimulus identity and the entries indicate the coordinates of the stimulus along the first (dim_0) and second (dim_1) dimensions of the embedding.
