---
title: "Characterizing Individual Differences Along a Continuous Manifold"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Characterizing Individual Differences Along a Continuous Manifold}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 6
)
```

## Introduction

The other vignettes in this package mostly treat individual differences in representation as a question of *clustering*: do participants fall into a small number of qualitatively distinct groups (see `vignette("tripletTools")`'s clustering section, and `test_for_clusters()`/`test_cluster_stability()`)? That's a natural first question, but it assumes the answer is discrete. Sometimes it isn't -- individual differences can instead vary smoothly along a *continuum*, with no natural dividing line between participants at one end and the other.

This vignette walks through that alternative using a real dataset: triplet similarity judgments on 58 color patches spanning color space, from 46 participants, collected by Clementine Zimnicki and described in a poster presented at the Annual Meeting of the Cognitive Science Society (Zimnicki et al.; <https://escholarship.org/uc/item/8778p3t3>). Participants judged which of two option patches was more similar in color to a reference patch, with no further instruction about what "similar" should mean. It covers:

- `generalized_procrustes()` -- aligning many embeddings into one shared consensus frame
- `smooth_embedding_trajectory()` -- a kernel-weighted average embedding at points along a continuous 1-D axis of individual variation
- `plot_3d_trajectory()`/`plot_2d_trajectory()` -- visualizing how the representation changes along that axis
- `test_dominant_dimension()` -- a direct statistical test for a single dominant axis, where Horn's parallel analysis is known to be conservative
- `loo_trajectory_accuracy()` -- testing whether manifold position predicts held-out judgments, without the data leakage a naive version of this test would have

The dataset is bundled with the package as `color_triplets` (raw judgments), `color_emb_ind`/`color_emb_group` (precomputed 3-D embeddings, fit the same way as in `vignette("embedding_vignette")`), and `color_lab` (the patches' own CIE LAB coordinates, for comparison).

```{r load}
library(tripletTools)
```

---

## The stimuli

The 58 patches span color space fairly evenly in CIE LAB coordinates. Their hex codes double as both item identifiers and plotting colors throughout this vignette. `color_lab` and the embedding objects (`color_emb_group`/`color_emb_ind`) aren't guaranteed to list the 58 patches in the same row order (confirmed they don't, here), so we fix one canonical order up front -- matching the embeddings, since that's what most of this vignette plots -- and derive `pcols` from it, rather than re-deriving row order (and risking a mismatch) in every later chunk:

```{r stimuli-plot, fig.width = 8, fig.height = 4.5, fig.cap="The 58 color patches plotted in their true CIE LAB coordinates (lightness vs. a, and b vs. a)."}
pcols <- rownames(color_emb_group) # canonical item order used throughout
lab_matched <- color_lab[pcols, ]  # color_lab re-indexed to match it

par(mfrow = c(1, 2), mar = c(3, 3, 2, 1))
plot(lab_matched[, c("l", "a")], pch = 22, bg = pcols, col = "gray",
     xlab = "L", ylab = "a", asp = 1, cex = 1.6)
plot(lab_matched[, c("b", "a")], pch = 22, bg = pcols, col = "gray",
     xlab = "b", ylab = "a", asp = 1, cex = 1.6)
```

## Does the group embedding recover real color space?

```{r align, fig.width = 8, fig.height = 8, fig.cap="True CIE LAB coordinates (top) vs. the group triplet embedding aligned to them via Procrustes (bottom)."}
pr <- vegan::procrustes(lab_matched, color_emb_group, symmetric = TRUE)
aligned <- pr$Yrot
rownames(aligned) <- pcols

cat("Procrustes correlation:", round(sqrt(1 - pr$ss), 3), "\n")

par(mfrow = c(2, 2), mar = c(3, 3, 2, 1))
plot(lab_matched[, c("l", "a")], pch = 22, bg = pcols, col = "gray",
     xlab = "L", ylab = "a", asp = 1, cex = 1.6, main = "True LAB")
plot(lab_matched[, c("b", "a")], pch = 22, bg = pcols, col = "gray",
     xlab = "b", ylab = "a", asp = 1, cex = 1.6, main = "True LAB")
plot(aligned[, 1:2], pch = 22, bg = pcols, col = "gray", cex = 1.6,
     xlab = "dim 1", ylab = "dim 2", main = "Aligned group embedding")
plot(aligned[, c(3, 2)], pch = 22, bg = pcols, col = "gray", cex = 1.6,
     xlab = "dim 3", ylab = "dim 2", main = "Aligned group embedding")
```

The two spaces are correlated overall (Procrustes correlation ~0.87) -- the group embedding does roughly recover real color space. But the recovered arrangement is visibly *not* the regular grid that true LAB space is: patches bow into a warped, roughly ring-shaped arrangement rather than staying evenly spaced. That's already a hint that participants' judgments are organized by something closer to color *category* than by precise perceptual distance -- a useful fact to hold onto, since it's exactly the kind of structure the trajectory analysis below will unpack further.

## Is there discrete cluster structure across participants?

Before assuming individual differences are continuous, it's worth checking whether they're actually discrete. `get.rep.dist()` gives a participant-by-participant representational distance matrix from the individual embeddings; `test_for_clusters()` checks both whether there's any real cluster structure at all (Hopkins statistic) and, if so, how many clusters (BIC):

```{r repdist}
repdist <- get.rep.dist(color_emb_ind)
```

```{r tfc, eval = requireNamespace("mclust", quietly = TRUE)}
test_for_clusters(repdist, max_clusters = 5, seed = 3)
```

The Hopkins statistic is non-significant and the BIC search prefers a single component (`best_g = 1`) -- no evidence of discrete subgroups. That rules out the clustering approach used elsewhere in this package and motivates treating individual differences as continuous instead.

## A continuous axis of individual differences

`estimate_intrinsic_dimension()` asks how many real dimensions of variation exist among participants' embeddings (as opposed to how many dimensions the color embeddings themselves have):

```{r intrinsic-dim}
estimate_intrinsic_dimension(repdist, seed = 3)
```

This comes back "forced to 1" -- Horn's parallel analysis has to beat a shuffled version of the leading dimension's own (already dominant) variance, which is an unusually hard bar to clear, and is a known false-negative pattern for exactly this case (see `?estimate_intrinsic_dimension`). But two further pieces of evidence point the same direction: the leading eigenvalue is about 5x the typical size of the rest, and -- more directly -- `test_dominant_dimension()`'s simulation-based significance test (run automatically above, since `test_dominance = TRUE` by default) rejects the no-structure null outright (p = 0.001). Unlike the eigenvalue-magnitude comparison Horn's test makes, this test compares a scale-invariant statistic (proportion of variance explained) against a null simulated at equal variance across candidate dimensions, which sidesteps the "beat a shuffled version of itself" problem -- see `?test_dominant_dimension` for why, and for a worked example where this same test does *not* reach significance despite a similarly large eigenvalue ratio (smaller sample sizes have less power to detect the same effect size). Combined with the non-significant Hopkins result above (no evidence of multiple discrete components either), this is solid evidence for one real, continuous dimension of individual variation -- not zero dimensions of structure.

`cmdscale()` gives each participant's own position along that one dimension:

```{r cmdscale, fig.cap="Distribution of participants' positions along the single estimated axis of representational variation."}
pos <- cmdscale(repdist, 1)
hist(pos, breaks = 15, xlab = "Manifold position", main = "")
```

Positions spread fairly evenly across the axis, with no obvious gap separating two groups -- consistent with a genuine continuum rather than two clusters that `test_for_clusters()` simply failed to detect.

## A smoothed trajectory along the manifold

`smooth_embedding_trajectory()` takes each participant's embedding and their position along this axis, and computes a kernel-weighted average embedding at a grid of query points spanning the axis. Internally, it first runs `generalized_procrustes()` to align every participant's embedding into one shared consensus frame (via iterative GPA, Gower 1975) -- without that step, averaging embeddings that are each in their own arbitrary rotation would wash out any real structure.

```{r traj}
traj <- smooth_embedding_trajectory(color_emb_ind, pos, n_query = 16)
names(traj)
```

`traj$effective_n` (Kish's effective sample size) tells us how many participants are really contributing to each query point -- useful for flagging the boundary regions, which are necessarily driven by fewer people:

```{r effn}
range(traj$effective_n)
```

## Visualizing the trajectory

`plot_3d_trajectory()` draws one query point at a time, with a gray strip along the bottom showing `effective_n` across the whole axis and a red marker showing which point is currently displayed. Plotting all 16 query points in sequence shows the representation changing smoothly:

```{r traj-grid, fig.width = 8, fig.height = 8, eval = requireNamespace("scatterplot3d", quietly = TRUE), fig.cap="The group representation at 16 points along the manifold, from one end (top-left) to the other (bottom-right)."}
par(mfrow = c(4, 4), mar = c(1, 1, 1, 1), oma = c(1, 1, 1, 1))
for (i in 1:16) {
  plot_3d_trajectory(traj, pos, loc = i, pcols = pcols, bcols = "gray")
}
```

The shift is striking: at one end of the manifold the patches collapse toward a nearly one-dimensional line (consistent with a strategy driven mostly by lightness), while at the other end they spread into a ring -- the same warped, category-like hue structure seen in the group embedding above, but now attributable to specific participants rather than smeared across everyone by averaging. `plot_2d_trajectory()` provides the same strip/marker/dot visualization for 2-D embeddings, using base `graphics::plot()` instead of `scatterplot3d` (not shown here, since these embeddings are 3-D).

## Does manifold position predict individual judgments?

If this axis reflects a real difference in how participants represent color, then each participant's own held-out judgments should be better predicted by the embedding from *their* end of the manifold than by the embedding from the opposite end.

It's tempting to test this by reusing the `traj` object already computed above: for each participant, evaluate `get.hoacc()` at the query point nearest their own position, and see whether that beats the opposite end. **This approach has a real, confirmed data-leakage problem, and should not be used.** `smooth_embedding_trajectory()`'s kernel weighting gives no participant's weight at a query point ever exactly zero, so every participant's own already-fitted embedding contributes -- with real weight -- to the very query-point embeddings later evaluated against their own held-out trials. A participant's own embedding unsurprisingly predicts their own behavior reasonably well regardless of whether any real shared structure exists at all, so this setup cannot distinguish "real structure shared with actual neighbors" from "trivial self-prediction via self-contribution to the averaged embedding." This was not a hypothetical concern: on synthetic data built from participants with fully independent, unrelated true embeddings (zero real shared structure, by construction), this naive approach produced a strong, "significant"-looking crossover (R<sup>2</sup> = 0.27, p = 0.003) purely from this mechanism (see `?smooth_embedding_trajectory`'s own warning section, and `?loo_trajectory_accuracy` for the full story).

`loo_trajectory_accuracy()` does this correctly: for every participant, it drops them entirely -- from the distance matrix before `cmdscale()` computes the axis, and from the embedding list before the trajectory is refit on the remaining participants -- and only then evaluates their held-out trials against the resulting, fully out-of-sample trajectory.

```{r loo-compute}
loo_res <- loo_trajectory_accuracy(repdist, color_emb_ind, color_triplets, n_query = 16, verbose = FALSE)
```

```{r crossover-plot, fig.width = 7, fig.height = 6, fig.cap="Leave-one-out held-out prediction accuracy from the two end-of-manifold embeddings, plotted against each participant's own (full-sample) manifold position."}
fsp <- loo_res$full_sample_position
acc <- loo_res$accuracy

plot(0, .5, type = "n", xlim = range(fsp), ylim = c(min(acc) - .05, max(acc) + .05),
     xlab = "Manifold position", ylab = "Accuracy on held-out validation trials")
for (i in seq_along(fsp)) lines(c(fsp[i], fsp[i]), c(acc[i, 1], acc[i, 16]), col = "gray")

points(fsp, acc[, 1], pch = 16, col = "steelblue")
abline(lm(acc[, 1] ~ fsp), col = "steelblue")

points(fsp, acc[, 16], pch = 16, col = "firebrick")
abline(lm(acc[, 16] ~ fsp), col = "firebrick")

abline(h = .5, lty = 2)
legend("bottomleft", legend = c("Embedding at left end", "Embedding at right end"),
       col = c("steelblue", "firebrick"), pch = 16, bty = "n")
```

The crossover survives: the left-end embedding predicts best for participants near the left end of the manifold, the right-end embedding predicts best for everyone else, and which one wins tracks manifold position almost perfectly (R<sup>2</sup> = `r round(summary(lm(I(acc[,1]-acc[,16]) ~ fsp))$r.squared, 2)`, from a simple linear regression of the accuracy difference on position -- barely smaller than what the leakage-prone naive version would have reported). This is a genuine, strong individual-differences effect -- in a domain (color similarity) where that kind of heterogeneity might not be expected at all -- and unlike the naive version, this result is actually trustworthy: it was computed with no participant ever contributing to the embedding used to predict them.

## Summary

| Question | Function(s) |
|---|---|
| Does the embedding recover a known reference space? | `vegan::procrustes()` |
| Is there discrete cluster structure across participants? | `test_for_clusters()` |
| Is there a real, continuous axis of individual variation? | `estimate_intrinsic_dimension()`, `test_dominant_dimension()`, `cmdscale()` |
| How does representation change smoothly along that axis? | `generalized_procrustes()`, `smooth_embedding_trajectory()` |
| How do I visualize that change? | `plot_3d_trajectory()`, `plot_2d_trajectory()` |
| Does manifold position predict individual judgments? | `loo_trajectory_accuracy()` (not `get.hoacc()` on `smooth_embedding_trajectory()`'s output directly -- see above) |

See `vignette("tripletTools")` for the discrete-clustering alternative (`test_for_clusters()`, `test_cluster_stability()`, hierarchical clustering), and `vignette("comparing_embeddings_vignette")` for more on Procrustes alignment and classifier-based evaluation.
