---
title: "Characterizing Individual Differences Along a Continuous Manifold"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Characterizing Individual Differences Along a Continuous Manifold}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 6
)
```

## Introduction

The other vignettes in this package mostly treat individual differences in representation as a question of *clustering*: do participants fall into a small number of qualitatively distinct groups (see `vignette("tripletTools")`'s clustering section, and `test_for_clusters()`/`test_cluster_stability()`)? That's a natural first question, but it assumes the answer is discrete. Sometimes it isn't -- individual differences can instead vary smoothly along a *continuum*, with no natural dividing line between participants at one end and the other.

This vignette walks through that alternative using a real dataset: triplet similarity judgments on 58 color patches spanning color space, from 46 participants, collected by Clementine Zimnicki and described in a poster presented at the Annual Meeting of the Cognitive Science Society (Zimnicki et al.; <https://escholarship.org/uc/item/8778p3t3>). Participants judged which of two option patches was more similar in color to a reference patch, with no further instruction about what "similar" should mean. It covers:

- `generalized_procrustes()` -- aligning many embeddings into one shared consensus frame
- `smooth_embedding_trajectory()` -- a kernel-weighted average embedding at points along a continuous 1-D axis of individual variation
- `plot_3d_trajectory()`/`plot_2d_trajectory()` -- visualizing how the representation changes along that axis

The dataset is bundled with the package as `color_triplets` (raw judgments), `color_emb_ind`/`color_emb_group` (precomputed 3-D embeddings, fit the same way as in `vignette("embedding_vignette")`), and `color_lab` (the patches' own CIE LAB coordinates, for comparison).

```{r load}
library(tripletTools)
```

---

## The stimuli

The 58 patches span color space fairly evenly in CIE LAB coordinates. Their hex codes double as both item identifiers and plotting colors throughout this vignette. `color_lab` and the embedding objects (`color_emb_group`/`color_emb_ind`) aren't guaranteed to list the 58 patches in the same row order (confirmed they don't, here), so we fix one canonical order up front -- matching the embeddings, since that's what most of this vignette plots -- and derive `pcols` from it, rather than re-deriving row order (and risking a mismatch) in every later chunk:

```{r stimuli-plot, fig.width = 8, fig.height = 4.5, fig.cap="The 58 color patches plotted in their true CIE LAB coordinates (lightness vs. a, and b vs. a)."}
pcols <- rownames(color_emb_group) # canonical item order used throughout
lab_matched <- color_lab[pcols, ]  # color_lab re-indexed to match it

par(mfrow = c(1, 2), mar = c(3, 3, 2, 1))
plot(lab_matched[, c("l", "a")], pch = 22, bg = pcols, col = "gray",
     xlab = "L", ylab = "a", asp = 1, cex = 1.6)
plot(lab_matched[, c("b", "a")], pch = 22, bg = pcols, col = "gray",
     xlab = "b", ylab = "a", asp = 1, cex = 1.6)
```

## Does the group embedding recover real color space?

```{r align, fig.width = 8, fig.height = 8, fig.cap="True CIE LAB coordinates (top) vs. the group triplet embedding aligned to them via Procrustes (bottom)."}
pr <- vegan::procrustes(lab_matched, color_emb_group, symmetric = TRUE)
aligned <- pr$Yrot
rownames(aligned) <- pcols

cat("Procrustes correlation:", round(sqrt(1 - pr$ss), 3), "\n")

par(mfrow = c(2, 2), mar = c(3, 3, 2, 1))
plot(lab_matched[, c("l", "a")], pch = 22, bg = pcols, col = "gray",
     xlab = "L", ylab = "a", asp = 1, cex = 1.6, main = "True LAB")
plot(lab_matched[, c("b", "a")], pch = 22, bg = pcols, col = "gray",
     xlab = "b", ylab = "a", asp = 1, cex = 1.6, main = "True LAB")
plot(aligned[, 1:2], pch = 22, bg = pcols, col = "gray", cex = 1.6,
     xlab = "dim 1", ylab = "dim 2", main = "Aligned group embedding")
plot(aligned[, c(3, 2)], pch = 22, bg = pcols, col = "gray", cex = 1.6,
     xlab = "dim 3", ylab = "dim 2", main = "Aligned group embedding")
```

The two spaces are correlated overall (Procrustes correlation ~0.87) -- the group embedding does roughly recover real color space. But the recovered arrangement is visibly *not* the regular grid that true LAB space is: patches bow into a warped, roughly ring-shaped arrangement rather than staying evenly spaced. That's already a hint that participants' judgments are organized by something closer to color *category* than by precise perceptual distance -- a useful fact to hold onto, since it's exactly the kind of structure the trajectory analysis below will unpack further.

## Is there discrete cluster structure across participants?

Before assuming individual differences are continuous, it's worth checking whether they're actually discrete. `get.rep.dist()` gives a participant-by-participant representational distance matrix from the individual embeddings; `test_for_clusters()` checks both whether there's any real cluster structure at all (Hopkins statistic) and, if so, how many clusters (BIC):

```{r repdist}
repdist <- get.rep.dist(color_emb_ind)
```

```{r tfc, eval = requireNamespace("mclust", quietly = TRUE)}
test_for_clusters(repdist, max_clusters = 5, seed = 3)
```

The Hopkins statistic is non-significant and the BIC search prefers a single component (`best_g = 1`) -- no evidence of discrete subgroups. That rules out the clustering approach used elsewhere in this package and motivates treating individual differences as continuous instead.

## A continuous axis of individual differences

`estimate_intrinsic_dimension()` asks how many real dimensions of variation exist among participants' embeddings (as opposed to how many dimensions the color embeddings themselves have):

```{r intrinsic-dim}
estimate_intrinsic_dimension(repdist)
```

This comes back "forced to 1" -- but with the leading eigenvalue several times the size of the rest, which is exactly the known false-negative pattern documented for this function (see `?estimate_intrinsic_dimension`): when nearly all the real signal sits in one dominant dimension, Horn's parallel analysis has to beat a shuffled version of that same dominant variance, which is an unusually hard bar to clear. Combined with the non-significant Hopkins result above (no evidence of multiple discrete components either), this is good evidence for one real, continuous dimension of individual variation -- not zero dimensions of structure.

`cmdscale()` gives each participant's own position along that one dimension:

```{r cmdscale, fig.cap="Distribution of participants' positions along the single estimated axis of representational variation."}
pos <- cmdscale(repdist, 1)
hist(pos, breaks = 15, xlab = "Manifold position", main = "")
```

Positions spread fairly evenly across the axis, with no obvious gap separating two groups -- consistent with a genuine continuum rather than two clusters that `test_for_clusters()` simply failed to detect.

## A smoothed trajectory along the manifold

`smooth_embedding_trajectory()` takes each participant's embedding and their position along this axis, and computes a kernel-weighted average embedding at a grid of query points spanning the axis. Internally, it first runs `generalized_procrustes()` to align every participant's embedding into one shared consensus frame (via iterative GPA, Gower 1975) -- without that step, averaging embeddings that are each in their own arbitrary rotation would wash out any real structure.

```{r traj}
traj <- smooth_embedding_trajectory(color_emb_ind, pos, n_query = 16)
names(traj)
```

`traj$effective_n` (Kish's effective sample size) tells us how many participants are really contributing to each query point -- useful for flagging the boundary regions, which are necessarily driven by fewer people:

```{r effn}
range(traj$effective_n)
```

## Visualizing the trajectory

`plot_3d_trajectory()` draws one query point at a time, with a gray strip along the bottom showing `effective_n` across the whole axis and a red marker showing which point is currently displayed. Plotting all 16 query points in sequence shows the representation changing smoothly:

```{r traj-grid, fig.width = 8, fig.height = 8, eval = requireNamespace("scatterplot3d", quietly = TRUE), fig.cap="The group representation at 16 points along the manifold, from one end (top-left) to the other (bottom-right)."}
par(mfrow = c(4, 4), mar = c(1, 1, 1, 1), oma = c(1, 1, 1, 1))
for (i in 1:16) {
  plot_3d_trajectory(traj, pos, loc = i, pcols = pcols, bcols = "gray")
}
```

The shift is striking: at one end of the manifold the patches collapse toward a nearly one-dimensional line (consistent with a strategy driven mostly by lightness), while at the other end they spread into a ring -- the same warped, category-like hue structure seen in the group embedding above, but now attributable to specific participants rather than smeared across everyone by averaging. `plot_2d_trajectory()` provides the same strip/marker/dot visualization for 2-D embeddings, using base `graphics::plot()` instead of `scatterplot3d` (not shown here, since these embeddings are 3-D).

## Does manifold position predict individual judgments?

If this axis reflects a real difference in how participants represent color, then each participant's own held-out judgments should be better predicted by the embedding from *their* end of the manifold than by the embedding from the opposite end. `get.hoacc()` gives, for one embedding, the proportion of a participant's held-out validation trials it predicts correctly:

```{r crossover-compute}
n <- length(color_triplets)
acc <- matrix(NA, n, 2)
for (i in seq_len(n)) {
  acc[i, 1] <- get.hoacc(traj$embeddings[[1]],  color_triplets[[i]], trialtype = "validation")
  acc[i, 2] <- get.hoacc(traj$embeddings[[16]], color_triplets[[i]], trialtype = "validation")
}
```

```{r crossover-plot, fig.width = 7, fig.height = 6, fig.cap="Held-out prediction accuracy from the two end-of-manifold embeddings, plotted against each participant's own manifold position."}
plot(0, .5, type = "n", xlim = range(pos), ylim = c(min(acc) - .05, max(acc) + .05),
     xlab = "Manifold position", ylab = "Accuracy on held-out validation trials")
for (i in seq_len(n)) lines(c(pos[i], pos[i]), c(acc[i, 1], acc[i, 2]), col = "gray")

points(pos, acc[, 1], pch = 16, col = "steelblue")
abline(lm(acc[, 1] ~ pos[, 1]), col = "steelblue")

points(pos, acc[, 2], pch = 16, col = "firebrick")
abline(lm(acc[, 2] ~ pos[, 1]), col = "firebrick")

abline(h = .5, lty = 2)
legend("bottomleft", legend = c("Embedding at left end", "Embedding at right end"),
       col = c("steelblue", "firebrick"), pch = 16, bty = "n")
```

A clean crossover: the left-end embedding predicts best for participants near the left end of the manifold, the right-end embedding predicts best for everyone else, and which one wins tracks manifold position almost perfectly (R<sup>2</sup> = `r round(summary(lm(I(acc[,1]-acc[,2]) ~ pos[,1]))$r.squared, 2)`, from a simple linear regression of the accuracy difference on position). This is a genuine, strong individual-differences effect -- in a domain (color similarity) where that kind of heterogeneity might not be expected at all.

## Summary

| Question | Function(s) |
|---|---|
| Does the embedding recover a known reference space? | `vegan::procrustes()` |
| Is there discrete cluster structure across participants? | `test_for_clusters()` |
| Is there a real, continuous axis of individual variation? | `estimate_intrinsic_dimension()`, `cmdscale()` |
| How does representation change smoothly along that axis? | `generalized_procrustes()`, `smooth_embedding_trajectory()` |
| How do I visualize that change? | `plot_3d_trajectory()`, `plot_2d_trajectory()` |
| Does manifold position predict individual judgments? | `get.hoacc()` |

See `vignette("tripletTools")` for the discrete-clustering alternative (`test_for_clusters()`, `test_cluster_stability()`, hierarchical clustering), and `vignette("comparing_embeddings_vignette")` for more on Procrustes alignment and classifier-based evaluation.
