---
title: "Getting Started with tripletTools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with tripletTools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 7
)
```

## What is a triplet task?

In a triplet (triadic comparison) task, a participant sees a *referent* item and two *option* items, and judges which option is more similar to the referent. From many such judgments across many items and participants, `tripletTools` computes a **similarity embedding**: a low-dimensional coordinate space in which items that are frequently judged as similar end up placed nearby.

This guide walks through the whole pipeline, in order: installation, one-time setup, fitting an embedding, and evaluating/visualizing the result. Every function used here is documented in more depth elsewhere — see [Where to go next] at the end.

## Installation

```{r install, eval = FALSE}
install.packages("remotes")
remotes::install_github("Knowledge-and-Concepts-Lab/tripletTools", build_vignettes = TRUE)
```

```{r load}
library(tripletTools)
```

## One-time setup: the Python embedding backend

Fitting a *new* embedding from triplet judgments runs inside a self-contained conda environment (PyTorch, NumPy, and friends), since that's where the actual optimization happens. Set it up once, ever, per machine:

```{r setup-env, eval = FALSE}
setup_python_env()
```

PyTorch is a sizeable download (300–800MB for a CPU build), so the first run takes a few minutes. You do **not** need to call this again in future R sessions — the environment is detected and activated automatically when the package loads. See `?setup_python_env` if conda isn't found on your system, or the "Computing Triplet Embeddings" vignette for GPU/CUDA setup.

**If you only want to analyze embeddings someone else already computed** (rather than fitting new ones from raw judgments), you can skip this step entirely — none of the analysis, evaluation, or comparison tools in this package touch Python.

## Explore the bundled example data

`tripletTools` ships a complete worked example: triplet judgments from 6 participants on 32 icon stimuli (faces and buildings), the icon images themselves, and a precomputed group embedding — enough to try the rest of this guide with no fitting required.

```{r explore-data}
# One participant's triplet judgments
str(icon_triplets[[1]])

# The 32 stimulus images
length(icon_pics)
```

## Visualize an embedding

`icon_emb_group` is a precomputed 3D group embedding of the 32 icons. `plot_pics()` plots each item's image at its embedding coordinates instead of a plain point, so you can see directly what the space captures:

```{r plot-group, fig.alt = "Icon stimuli plotted at their group-embedding coordinates"}
emb <- icon_emb_group
rownames(emb) <- emb$item

plot_pics(emb[, c("dim_0", "dim_1")], icon_pics,
          psize = 0.04,
          xlab  = "Dimension 1",
          ylab  = "Dimension 2",
          main  = "Group embedding of the icon stimuli")
```

Items placed close together were frequently judged as similar across participants.

## Evaluate an embedding

A natural question for any embedding is how well it predicts held-out judgments it wasn't fit on. `get.hoacc()` answers this directly:

```{r evaluate}
acc <- get.hoacc(emb[, c("dim_0", "dim_1", "dim_2")], icon_triplets[[1]])
cat("Hold-out accuracy (participant 1):", round(acc, 3), "\n")
```

## Fit your first embedding

With setup done (above), fitting a new embedding from triplet judgments is one function call:

```{r fit-embedding, eval = FALSE}
grp <- run_group_embedding_from_list(
  triplet_list = icon_triplets,
  d            = 3L,
  max_epochs   = 50000L
)

head(grp$embedding)
```

`d` (the number of dimensions) is a hyperparameter you'd normally choose via `estimate_dimensionality()` rather than guess — see the "Computing Triplet Embeddings" vignette for that and for fitting per-participant (not just group) embeddings.

## Where to go next

| Vignette | Covers |
|---|---|
| [tripletTools Overview](overview_vignette.html) | Loading data, data quality, inter-subject agreement, clustering participants |
| [Computing Triplet Embeddings](embedding_vignette.html) | The full embedding pipeline: dimensionality selection, individual embeddings, spherical/circular embeddings, parallel and HTCondor execution |
| [Comparing Triplet Embeddings to Alternative Representations](comparing_embeddings_vignette.html) | Comparing embeddings to language-model embeddings, evaluating them via classifier performance, estimating similarity from verbal-fluency data |
| [Read Triplet Data](read_data_vignette.html) | File formats and reading in data from a new study |
