---
title: "tripletTools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tripletTools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tripletTools)
```

This package contains a set of functions to aid analysis of data from _triadic comparisons_ or _triplet_ tasks. In triadic comparisons, participants must decide which of two _option_ items is most similar to a third _target_ item. From many such decisions one can compute a metric embedding of the stimulus items, such that items often chosen as similar to one another are nearby in the space. The suite of functions in this package allow the user to visualize and evaluate the quality of embeddings generated from triadic comparisons data.

## Data files and naming conventions

The functions make use of two kinds of data files: _triplet data files_, which contain information about each trial of a tradic comparison task, and _embedding files_, which contain the coordinates of each item in an embedding computed from triplet judgment data.

### Triplet data files

Triplet data files as `.csv` files generated by the software used to collect triplet judgment data. After a header specifying required column names, each row of the file indicates critical information for each trial of a triplet experiments. Typically data from all participants in a given study are included in a single triplet data file. 

The triplet data file should be a `.csv` file containing the following fields:

- _head_, _winner_, _loser_: Integer indices for each item appearing in the triplet
- _worker\_id_: Arbitrary identifier for each participant
- _rt_: Response time for the trial
- _Center_, _Left_, _Right_: Strings indicating the items appearing in the center (target item), left side (option 1) and right side (option 2).
- _Answer_: String indicating which option the participant chose.
- _sampleAlg_: ALgorithm used to sample the item: either random, validation, or check.
- _sampleSet_: Indicates whether the triplet was used to fit the embedding (train) or not (test).

Data in this format can be read into the current session using the function `get_combined(fname)` where `fname` is the path to the data file. This function returns a named list, where each element includes the triplet judgment data from a single subject, and elements are named by the subject identified. This package includes an example dataset in this format, `cfd_triplets`:

```{r}
head(cfd_triplets[[1]])
```
Here you can see the triplet judgment data for each trial for the first participant in the experiment. Participants viewed the trials in the same order they are listed in the matrix. To learn more about this dataset, try `help(cfd_triplets)`.

### Embedding data


